---
title: "plant-import-paper-r-markdown"
author: "pgfarsund"
date: "2023"
output: github_document
---

this markdown document contains code for the analysis and results already implemented 
in the manuscript for "the plant import paper" (in prep for Biological Invasions). in 
addition the these results, another script contain

```{r}
library(phyloseq)
library(tidyverse)
library(ggpubr)
```


load, clean, and manipulate phyloseq objects:
```{r}
### fungi

# load data:
plants = readRDS(file = "plants-phyloseq-object.RDS") # change the path to data's location on your computer.

# manipulate the sample data:
sam = data.frame(plants@sam_data) %>% 
  dplyr::select("sample","Location_name",            # there is a lot of data we don't need.
                "Specific_host",
                "Sample_ID") %>% 
  rename(c("container"="Location_name",              # rename columns for simplicity.
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number.
         sample=sub("_ITS_S2_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID.
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot.
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))) %>%
  mutate(interval = c(rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10), 
                      rep("double", times=10),
                      rep("single", times=10))); str(sam) 
sample_data(plants) = sample_data(sam); rm(sam); plants

# for this study we're only interested in vascular plants. our data set contains
# some bryophytes and green algae, additionally we're not interested in Taxus sp. 
# or Thuja sp., because these were ornamentals in the sampled soil. 
# remove non-focal taxa:
plants=subset_taxa(plants, phylum=="Streptophyta")
plants=subset_taxa(plants, class!="Chlorophyceae")
plants=subset_taxa(plants, order!="Splachnales")
plants=subset_taxa(plants, order!="Bryales")
plants=subset_taxa(plants, order!="Dicranales")
plants=subset_taxa(plants, order!="Funariales")
plants=subset_taxa(plants, order!="Hypnales")
plants=subset_taxa(plants, order!="Klebsormidiaceae")
plants=subset_taxa(plants, order!="Pottiales")
plants=subset_taxa(plants, order!="Pseudoditrichales")
plants=subset_taxa(plants, species!="Thuja_sp.")
plants=subset_taxa(plants, species!="Thuja_occidentalis")
plants=subset_taxa(plants, species!="Taxus_sp.")
plants=subset_taxa(plants, species!="Taxus_canadensis"); plants



### fungi

# load data:
fungi = readRDS(file = "fungi-phyloseq-object.RDS") # change the path to data's location on your computer
fungi = subset_samples(fungi, negative_control=="FALSE")

# manipulate the sample data:
sam = data.frame(fungi@sam_data) %>% 
  dplyr::select("sample","Location_name",            # there is a lot of data we don't need
                "Specific_host",
                "Sample_ID") %>%
  rename(c("container"="Location_name",              # rename columns for simplicity
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number
         sample=sub("_fITS7_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))) %>%
  mutate(interval = c(rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10), 
                      rep("double", times=10),
                      rep("single", times=10)))
sample_data(fungi) = sample_data(sam); rm(sam); fungi
```


##### RESULTS

### contaminant plants and fungi ### (Fig. 1)

we identified many contaminant plants and fungi that were native, known aliens, 
and potential alien species to Norway. we've grouped these species as "Native", 
"Alien", "Swe., Fin., Den." (i.e. native to Norway's neighboring countries), 
(native to) "Europe", and (native to) "Non-Europe". we will make a stacked bar
chart to show the relative frequencies of these different categories:

```{r}

# build the data frame:
df = data.frame("freq" = c(30.30, 45.45, 0, 12.12, 12.12, 0, # start with plant frequencies
                           39.53, 0.81, 22.32, 22.89, 10.88, 3.57), # then fungal frequencies
                "group" = rep(c("Plants",
                                "Fungi"), each=6),
                "cats" = c(rep(c("Native",
                                 "Alien",
                                 "Swe., Fin., Den.",
                                 "Europe", 
                                 "Non-Europe",
                                 "NA"), times=2)))

# make Figure 1 in the manuscript:
fig.1 = df %>%
ggplot(aes(x=factor(group, 
                    level=c("Plants",
                            "Fungi")), # order the groups on the x-axis
           y=freq, 
           fill=factor(cats, 
                       levels = c("Native", # we want the stacked bars to have this specific order
                                  "Alien", 
                                  "Swe., Fin., Den.",
                                  "Europe", 
                                  "Non-Europe", 
                                  "NA")),
           color=factor(cats, 
                        levels = c("Native", 
                                   "Alien", 
                                   "Swe., Fin., Den.",
                                   "Europe", 
                                   "Non-Europe", 
                                   "NA")))) + 
  geom_col(position = "fill", 
           width = 0.55) + 
  theme_classic(base_size = 15) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", # some nice colors to fill the bars
                               "#F0E442", "#CC79A7", "#999999")) +
  scale_color_manual(values = rep("black", times = 6)) +
  theme(legend.position = "right") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.title = element_blank()) +
  annotate(geom = "text", 
           label = c("30.3 %", 
                     "45.5 %",
                     "12.1 %",
                     "12.1 %", 
                     "39.5 %",
                     "0.8 %",
                     "22.3 %", 
                     "22.9 %", 
                     "10.9 %",
                     "3.6 %"),
           x = rep(c(1,2,2.45,2,2.45), # we want "0.8 %" and "3.6 %" outside the actual bar, which makes this a bit tricky
                   times = c(4,1,1,3,1)), 
           y = c(0.84, 0.45, 0.18, 0.06, # plant percent labels
                 0.8, 0.6, 0.5, 0.26, 0.1, 0.02), # fungi percent labels
           size = 4, color = "black", fontface = 1) + 
  annotate(geom = "segment", 
           x = rep(2.25,
                   times=2), 
           xend = rep(2.315, 
                      times=2), 
           y = c(0.60,
                 0.02), 
           yend = c(0.60,
                    0.02)) + 
  scale_y_continuous(labels = scales::percent); fig.1 # save as .pdf landscape 5 x 6

```


### metabarcoding vs. germination ### (Fig. 2)

we identified 33 plant species with metabarcoding and 30 with germination. five 
species were identified by both methods, adding up to 58 contaminant plant species
identified all together. we also checked whether identified plants were registered 
as products in the online assortments of Noviflora and Plantasjen in the fall of 
2021 (the year of sampling). we will make two stacked bar charts to show how many 
species fall into each category, while also showing how many species were identified
with metabarcoding, germination, or both: 

```{r}
df2 = read.csv(file = "C:/Users/pgfar/Dropbox/Whats_in_the_pot/plant-import-paper/fig_2_dataset.csv",
               sep = ";"); str(df2) # change path 

my_colors = c("#CC79A7",
              "#009E73",
              "#E69F00")

a = df2 %>% 
  mutate(prev_det=sub("TRUE", "Yes", prev_det)) %>% 
  mutate(prev_det=sub("FALSE", "No", prev_det)) %>% 
  ggplot(aes(x=factor(prev_det, 
                             levels = c("Yes", 
                                        "No")),
                color=factor(id_method,
                             levels = c("Metabarcoding",
                                        "Germination",
                                        "Both")),
                fill=factor(id_method,
                            levels = c("Metabarcoding",
                                       "Germination",
                                       "Both")))) + 
  geom_bar(position = "stack",
           width = 0.55) +
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = rep("black", times=3)) +
  theme_classic(base_size = 15) + 
  xlab("Previously identified?") + 
  ylab("Number of plant species") + 
  labs(fill = "Identification\nmethod", 
       color = "Identification\nmethod", 
       subtitle = "a.") +
  theme(legend.title = element_text(size = 13)) +
  scale_y_continuous(breaks = seq(0, 45, 10), 
                     limits = c(0, 45)); a

b = df2 %>% 
  mutate(ornamental=sub("TRUE", "Yes", ornamental)) %>% 
  mutate(ornamental=sub("FALSE", "No", ornamental)) %>% 
  ggplot(aes(x=factor(ornamental, 
                             levels = c("Yes", 
                                        "No")),
                color=factor(id_method,
                           levels = c("Metabarcoding",
                                      "Germination",
                                      "Both")),
                fill=factor(id_method,
                           levels = c("Metabarcoding",
                                      "Germination",
                                      "Both")))) + 
  geom_bar(position = "stack",
           width = 0.55) +
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = rep("black", times=3)) +
  theme_classic(base_size = 15) + 
  xlab("Ornamental?") + 
  ylab(" ") + 
  labs(fill = "Identification\nmethod", 
       color = "Identification\nmethod", 
       subtitle = "b.") +
  theme(legend.title = element_text(size = 13)) +
  scale_y_continuous(breaks = seq(0, 55, 10), 
                     limits = c(0, 45)); b

fig.2 = ggarrange(a, b, 
                    common.legend = TRUE,
                    legend = "right", 
                    nrow = 1); fig.2

```


### effect of sampling strategy on contaminant species richness ### (Fig. 3)

we want to investigate the effect of sampling effort on the observed species 
richness of contaminant plants and fungi identified by metabarcoding. we'll do
this in two steps:

1. we want to know whether a larger volume of soil per sample had an effect on 
the observed species richness per pot. unfortunately, all samples consited of 
approximately 25 grams. we're therefore going to simulate doubling the soil sample 
volume by merging samples that were taken from the same same pot, prior to 
normalizing the read counts across all samples:

```{r}

# we're going to merge samples from the same pots. to do this, we first create a 
# new phyloseq object with only the samples we're going to merge: 
twice = subset_samples(plants, interval == "double"); twice
twice.m = merge_samples(twice, group = "pot_id"); twice.m 

# reassign sample data in twice.m:
twice.m@sam_data$pot_id = rownames(twice.m@sam_data) # fix pot_id variable
twice.m@sam_data$pot_id = sub("p","",twice.m@sam_data$pot_id) 
twice.m@sam_data$pot_id = as.numeric(twice.m@sam_data$pot_id) # this will let us sort by pot_id
twice.m.sam = data.frame(twice.m@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id))

# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(sample_id!="1_2") %>%
  filter(sample_id!="2_2") %>%
  filter(sample_id!="3_2") %>%
  filter(sample_id!="4_2") %>%
  filter(sample_id!="5_2")

# and reassign variables: 
twice.m.sam$host = sam$host
twice.m.sam$container = sam$container
twice.m.sam$interval = sam$interval

# convert til sample_data() object:
twice.m.sam = sample_data(twice.m.sam); sample_names(twice.m.sam)

# and reassign to twice.m phyloseq object: 
sample_data(twice.m) = twice.m.sam; twice.m; head(twice.m@sam_data)


# now we want to get the merged samples back in the original dataset. we first
# create a new dataset with only samples from pots sampled once: 
once = subset_samples(plants, interval == "single"); once

# and now we can merge the two phyloseq objects:
plants2 = merge_phyloseq(twice.m, once); plants2

sort(sample_sums(plants2))
f2rar = rarefy_even_depth(plants2, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE, 
                          sample.size = 14215); plants2; f2rar

# now we're ready to have a look at observed species richness:
# make a dataframe for ease of testing and plotting:
obs.df = data.frame(cbind(estimate_richness(f2rar, 
                                            measures = "Observed"), 
                          sample_data(f2rar))); str(obs.df)

hist(obs.df$Observed); shapiro.test(obs.df$Observed) # not statistically significantly different from a normal distribution...
var.test(Observed ~ interval, obs.df) # the variance in the two groups is not significantly different from each other
mod1=t.test(Observed ~ interval, obs.df, var.equal=TRUE)
si=obs.df %>% 
  filter(interval=="single"); sd(si$Observed)
db=obs.df %>% 
  filter(interval=="double"); sd(db$Observed)

# make a nice boxplot to go with the results:
fig3.a.p=obs.df %>% 
  ggplot(aes(x=interval, y=Observed, 
             fill=interval)) + 
  geom_boxplot(alpha = 0.6) + 
  scale_fill_manual(values = c("chartreuse2", 
                               "steelblue1")) +
  theme_classic() + 
  ylab("Observed species richness") +
  scale_x_discrete(labels = c("Double", 
                              "Single")) +
  theme(legend.position = "none") +
  labs(title = "A.", 
       subtitle = "Plants") +
  xlab("Soil sample volume") +
  annotate(geom = "text", 
           x=1.5, 
           y=15, 
           label = paste0("P = ",
                          round(mod1$p.value, 
                                digits = 2)), # annotate rho and p-value
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0); fig3.a.p

```

to go along with the results about pots sampled twice, we're interested in the
overall recapture rate of the species that were identified in pots sampled twice.
i.e., were species identified consistently in both samples from the pots they
were identified in? or rather, was two samples (or increased volume) necessary
for identifying certain species in many pots?

!!!THIS CHUNK NEEDS CLEANING!!!
```{r}

# !!!THIS CHUNK NEEDS CLEANING!!!
#! !!THIS CHUNK NEEDS CLEANING!!!
#!! !THIS CHUNK NEEDS CLEANING!!!

x = dp

pf = subset_samples(x, turn=="first")

pf@sam_data$row = paste0(pf@sam_data$pot_id,
                        "_", 
                        pf@sam_data$turn); head(pf@sam_data)

psec = subset_samples(x, turn=="second")
psec@sam_data$row = paste0(psec@sam_data$pot_id,
                           "_", 
                           psec@sam_data$turn); head(psec@sam_data)

x = merge_phyloseq(pf, psec); x
sample_names(x) = x@sam_data$row

colnames(x@otu_table) = data.frame(x@tax_table)[,6]

df = data.frame(x@otu_table) %>% 
  mutate_if(is.numeric, ~1 * (. > 0))

df.t = data.frame(t(df))

first = data.frame(df.t) %>% 
  select(matches("_first"))
colnames(first) = sub("X", "", colnames(first))
colnames(first) = sub("_first", "", colnames(first))

second = data.frame(df.t) %>% 
  select(matches("_second"))
colnames(second) = sub("X", "", colnames(second))
colnames(second) = sub("_second", "", colnames(second))

test = data.frame(first+second)
colnames(test) = sub("X", "", colnames(test))
#



# this is where you find the overall recapture rate of 70.3 % 
# 
table(test==0) # 175 detections in total
table(test==1) # 123 = 1
table(test==2) # 52 = 2

c = test[rownames(test) %in% "Erigeron_canadensis", ]
table(c==0) 
table(c==1) 
table(c==2)

c = test[rownames(test) %in% "Acer_pseudoplatanus", ]
table(c==0) 
table(c==1) 
table(c==2)
#




ppp = full_join(pf, psec)

recapture = ppp %>% 
  distinct(species); recapture



```

2. next we want to calculate the fold increase in species richness per pot when
increasing from one to two samples. to do this we will only use data from pots
sampled twice. FI = fold increase, S = sample, Obs = observed species richness:
FI = (Obs S1 + Obs S2) / Obs S1

```{r}
twice = subset_samples(plants, interval == "double"); twice = prune_taxa(taxa_sums(twice)>0,
                                                                         twice); twice
first = subset_samples(twice, grepl(pattern = "_1", 
                                    x = twice@sam_data$sample_id)); first = prune_taxa(taxa_sums(first)>0,
                                                                                       first); first
first@sam_data$turn = "first"; head(first@sam_data) # add a variable to distinguish between first and merged samples
second = merge_samples(twice, group = "pot_id"); second 

# reassign sample data in second:
second@sam_data$pot_id = rownames(second@sam_data) # fix pot_id variable
second@sam_data$pot_id = sub("p","", second@sam_data$pot_id) 
second@sam_data$pot_id = as.numeric(second@sam_data$pot_id) # this will let us sort by pot_id
second.sam = data.frame(second@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id)); head(second.sam); tail(second.sam)



# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(grepl(pattern = "_2", 
               sam$sample_id))

# and reassign variables: 
second.sam$host = sam$host
second.sam$container = sam$container
second.sam$interval = sam$interval
second.sam$turn = "second"; # add a variable to distinguish between merged and first samples

# convert til sample_data() object:
second.sam = sample_data(second.sam); sample_names(second.sam)

# and reassign to twice.m phyloseq object: 
sample_data(second) = second.sam; twice.m; head(second@sam_data)


# now we can merge the two phyloseq objects in order to normalize them:
fiphy = merge_phyloseq(first, second); fiphy

# normalize the new phyloseq object:
sort(sample_sums(fiphy)) # we can normalize at the lowest read depth to keep all samples and only lose two taxa:
fiphyrar = rarefy_even_depth(fiphy, 
                             rngseed = 123, 
                             replace = FALSE, 
                             trimOTUs = TRUE, 
                             sample.size = 10378); fiphy; fiphyrar

# now we have to separate the new object into two dataframes before we combine
# them to calculate and plot fold increase: 
first = subset_samples(fiphyrar, turn=="first")
first.df = data.frame(cbind(estimate_richness(first, 
                                              measures = "Observed"),
                            first@sam_data)) %>% 
  rename("obs_first"="Observed",
         "pot_id_first"="pot_id",
         "turn_first"="turn") %>% 
  mutate(pot_id=pot_id_first); head(first.df)

second = subset_samples(fiphyrar, turn=="second")
second.df = data.frame(cbind(estimate_richness(second, 
                                              measures = "Observed"),
                            second@sam_data)) %>% 
  rename("obs_second"="Observed",
         "pot_id_second"="pot_id",
         "turn_second"="turn") %>% 
  mutate(pot_id=pot_id_second); head(second.df)

fidf = full_join(first.df, second.df, by = "pot_id") %>% 
  select("obs_first", "obs_second", 
         "pot_id", "pot_id_first", "pot_id_second", 
         "turn_first", "turn_second")

fidf = fidf %>% 
  mutate(foldincrease = obs_second/obs_first)
fig3.b.p=ggplot(fidf, aes(x=reorder(pot_id, -foldincrease), 
                 y=foldincrease,
                 fill="darkolivegreen1", color="black")) + 
  geom_col(width=1) +
  scale_fill_manual(values = "darkolivegreen1") +
  scale_color_manual(values = "black") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.ticks.x = element_blank()
      ) + 
  ylab("Fold increase") +
  xlab("Sampled pots") +
  scale_x_discrete(labels = paste0(rep(" ", times=20))) +
  labs(title = "B.",
       subtitle = "Plants") +
  annotate(geom = "text", 
           x=5, 
           y=2.5, 
           label = paste0("Median = ", 
                         round(median(fidf$foldincrease), 
                               digits = 2)), # calculate and annotate manually...
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0) +
  scale_y_continuous(breaks = seq(0, 3, 0.5), limits = c(0, 3)) + 
  coord_cartesian(ylim = c(0.9, 3)); fig3.b.p

# additionally, we will investigate whether there is a correlation between 
# the richness in the second sample and fold increase for each pot:
cormod = cor.test(fidf$foldincrease, fidf$obs_second, 
         method = "pearson")
fig3.c.p=ggplot(fidf, aes(x=obs_second, 
                 y=foldincrease,
                 fill="darkolivegreen1")) + 
  theme_classic() +
  theme(legend.position = "none", 
        ) +
  scale_color_manual(values = "black") + 
  scale_fill_manual(values = "darkolivegreen1") +
  geom_point(size=6, pch=21) + 
  geom_smooth(method = "lm", 
              color="black", 
              fill="grey90") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 4, 0.5), 
                    limits = c(0, 3.6)) +
  xlab("Richness in both samples") +
  ylab("\nFold increase") + 
  labs(title = "C.",
       subtitle = "Plants") +
  coord_cartesian(ylim = c(1,3)) + 
  annotate(geom = "text", 
           x=8.8, 
           y=2.8, 
           label = paste0("r = ", 
                          round(cormod$estimate, 
                                digits = 2),
                          "\nP = ",
                          round(cormod$p.value,
                                digits = 2)), 
           fontface = "italic", size=3, color="black", 
           hjust = 0); fig3.c.p


```

we now run the three previous chunks for fungi:

1. soil volume:

```{r}

# we're going to merge samples from the same pots. to do this, we first create a 
# new phyloseq object with only the samples we're going to merge: 
twice = subset_samples(fungi, interval == "double"); twice
twice.m = merge_samples(twice, group = "pot_id"); twice.m 

# reassign sample data in twice.m:
twice.m@sam_data$pot_id = rownames(twice.m@sam_data) # fix pot_id variable
twice.m@sam_data$pot_id = sub("p","",twice.m@sam_data$pot_id) 
twice.m@sam_data$pot_id = as.numeric(twice.m@sam_data$pot_id) # this will let us sort by pot_id
twice.m.sam = data.frame(twice.m@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id))

# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(sample_id!="1_2") %>%
  filter(sample_id!="2_2") %>%
  filter(sample_id!="3_2") %>%
  filter(sample_id!="4_2") %>%
  filter(sample_id!="5_2")

# and reassign variables: 
twice.m.sam$host = sam$host
twice.m.sam$container = sam$container
twice.m.sam$interval = sam$interval

# convert til sample_data() object:
twice.m.sam = sample_data(twice.m.sam); sample_names(twice.m.sam)

# and reassign to twice.m phyloseq object: 
sample_data(twice.m) = twice.m.sam; twice.m; head(twice.m@sam_data)


# now we want to get the merged samples back in the original dataset. we first
# create a new dataset with only samples from pots sampled once: 
once = subset_samples(fungi, interval == "single"); once

# and now we can merge the two phyloseq objects:
fungi2 = merge_phyloseq(twice.m, once); fungi2

sort(sample_sums(fungi2))
f2rar = rarefy_even_depth(fungi2, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE, 
                          sample.size = 237829); fungi2; f2rar

# now we're ready to have a look at observed species richness:
# make a dataframe for ease of testing and plotting:
obs.df = data.frame(cbind(estimate_richness(f2rar, 
                                            measures = "Observed"), 
                          sample_data(f2rar))); str(obs.df)

hist(obs.df$Observed); shapiro.test(obs.df$Observed) # not statistically significantly different from a normal distribution...
var.test(Observed ~ interval, obs.df) # the variance in the two groups is not significantly different from each other
mod1=t.test(Observed ~ interval, 
            obs.df, var.equal=TRUE)
si=obs.df %>% 
  filter(interval=="single"); sd(si$Observed)
db=obs.df %>% 
  filter(interval=="double"); sd(db$Observed)

# make a nice boxplot to go with the results:
fig3.a.f=obs.df %>% 
  ggplot(aes(x=interval, y=Observed, 
             fill=interval)) + 
  geom_boxplot(alpha = 0.6) + 
  scale_fill_manual(values = c("tan1", 
                               "gold1")) +
  theme_classic() + 
  ylab("Observed species richness") +
  scale_x_discrete(labels = c("Double", 
                              "Single")) +
  theme(legend.position = "none") +
  labs(title = " ", 
       subtitle = "Fungi") +
  xlab("Soil sample volume") +
  annotate(geom = "text", 
           x=1.5, 
           y=850, 
           label = paste0("P = <",
                          round(mod1$p.value, 
                                digits = 2),
                          ".001"), # annotate rho and p-value
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0); fig3.a.f

```

overall recapture rate:
!!!THIS CHUNK NEEDS CLEANING!!!
```{r}

#!!!THIS CHUNK NEEDS CLEANING!!!
#!!!THIS CHUNK NEEDS CLEANING!!!
#!!!THIS CHUNK NEEDS CLEANING!!!

x = df

pf = subset_samples(x, turn=="first")

pf@sam_data$row = paste0(pf@sam_data$pot_id,
                        "_", 
                        pf@sam_data$turn); head(pf@sam_data)

psec = subset_samples(x, turn=="second")
psec@sam_data$row = paste0(psec@sam_data$pot_id,
                           "_", 
                           psec@sam_data$turn); head(psec@sam_data)

x = merge_phyloseq(pf, psec); x
sample_names(x) = x@sam_data$row

colnames(x@otu_table) = data.frame(x@tax_table)[,7]

dff = data.frame(x@otu_table) %>% 
  mutate_if(is.numeric, ~1 * (. > 0))

df.t = data.frame(t(dff))

first = data.frame(df.t) %>% 
  select(matches("_first"))
colnames(first) = sub("X", "", colnames(first))
colnames(first) = sub("_first", "", colnames(first))

second = data.frame(df.t) %>% 
  select(matches("_second"))
colnames(second) = sub("X", "", colnames(second))
colnames(second) = sub("_second", "", colnames(second))

test = data.frame(first+second)
colnames(test) = sub("X", "", colnames(test))
#


table(test==0) # 14 498 detections total
table(test==1) # 8402 = 1 ##### 8402 / (8402+6096)
table(test==2) # 6096 = 2

c = test[rownames(test) %in% "Cryptostroma_corticale", ]
table(c==0) 
table(c==1) 
table(c==2)

c = test[rownames(test) %in% "Mutinus_ravenelii", ]
table(c==0) 
table(c==1) 
table(c==2)
#




fff = full_join(ff, fsec)

recapture = fff %>% 
  distinct(species); recapture



```


2. fold increase

```{r}
twice = subset_samples(fungi, interval == "double"); twice = prune_taxa(taxa_sums(twice)>0,
                                                                         twice); twice
first = subset_samples(twice, grepl(pattern = "_1", 
                                    x = twice@sam_data$sample_id)); first = prune_taxa(taxa_sums(first)>0,
                                                                                       first); first
first@sam_data$turn = "first"; head(first@sam_data) # add a variable to distinguish between first and merged samples
second = merge_samples(twice, group = "pot_id"); second 

# reassign sample data in second:
second@sam_data$pot_id = rownames(second@sam_data) # fix pot_id variable
second@sam_data$pot_id = sub("p","", second@sam_data$pot_id) 
second@sam_data$pot_id = as.numeric(second@sam_data$pot_id) # this will let us sort by pot_id
second.sam = data.frame(second@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id)); head(second.sam); tail(second.sam)



# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(grepl(pattern = "_2", 
               sam$sample_id))

# and reassign variables: 
second.sam$host = sam$host
second.sam$container = sam$container
second.sam$interval = sam$interval
second.sam$turn = "second"; # add a variable to distinguish between merged and first samples

# convert til sample_data() object:
second.sam = sample_data(second.sam); sample_names(second.sam)

# and reassign to twice.m phyloseq object: 
sample_data(second) = second.sam; twice.m; head(second@sam_data)


# now we can merge the two phyloseq objects in order to normalize them:
fiphy = merge_phyloseq(first, second); fiphy

# normalize the new phyloseq object:
sort(sample_sums(fiphy)) # we can normalize at the lowest read depth to keep all samples and only lose two taxa:
fiphyrar = rarefy_even_depth(fiphy, 
                             rngseed = 123, 
                             replace = FALSE, 
                             trimOTUs = TRUE, 
                             sample.size = 213969); fiphy; fiphyrar

# now we have to separate the new object into two dataframes before we combine
# them to calculate and plot fold increase: 
first = subset_samples(fiphyrar, turn=="first")
first.df = data.frame(cbind(estimate_richness(first, 
                                              measures = "Observed"),
                            first@sam_data)) %>% 
  rename("obs_first"="Observed",
         "pot_id_first"="pot_id",
         "turn_first"="turn") %>% 
  mutate(pot_id=pot_id_first); head(first.df)

second = subset_samples(fiphyrar, turn=="second")
second.df = data.frame(cbind(estimate_richness(second, 
                                              measures = "Observed"),
                            second@sam_data)) %>% 
  rename("obs_second"="Observed",
         "pot_id_second"="pot_id",
         "turn_second"="turn") %>% 
  mutate(pot_id=pot_id_second); head(second.df)

fidf = full_join(first.df, second.df, by = "pot_id") %>% 
  select("obs_first", "obs_second", 
         "pot_id", "pot_id_first", "pot_id_second", 
         "turn_first", "turn_second")

fidf = fidf %>% 
  mutate(foldincrease = obs_second/obs_first)

fig3.b.f=ggplot(fidf, aes(x=reorder(pot_id, -foldincrease), 
                 y=foldincrease,
                 fill="tan1", color="black")) + 
  geom_col(width=1) +
  scale_fill_manual(values = "tan1") +
  scale_color_manual(values = "black") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.ticks.x = element_blank()
      ) + 
  ylab("Fold increase") + 
  xlab("Sampled pots") +
  scale_x_discrete(labels = paste0(rep(" ", times=20))) +
  labs(title = " ",
       subtitle = "Fungi") +
  annotate(geom = "text", 
           x=5, 
           y=1.75, 
           label = paste0("Median = ", 
                         round(median(fidf$foldincrease), 
                               digits = 2)), # calculate and annotate manually...
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0) +
  scale_y_continuous(breaks = seq(0, 3, 0.5), limits = c(0, 3)) + 
  coord_cartesian(ylim = c(0.9, 3)); fig3.b.f

# additionally, we will investigate whether there is a correlation between 
# the richness in the second sample and fold increase for each pot:
cormod = cor.test(fidf$foldincrease, fidf$obs_second, 
         method = "pearson")
fig3.c.f=ggplot(fidf, aes(x=obs_second, 
                 y=foldincrease,
                 fill="tan1")) + 
  theme_classic() +
  theme(legend.position = "none", 
        ) +
  scale_color_manual(values = "black") + 
  scale_fill_manual(values = "tan1") +
  geom_point(size=6, pch=21) + 
  geom_smooth(method = "lm", 
              color="black", 
              fill="grey90") +
  #scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 4, 0.5), 
                    limits = c(0, 3.6)) +
  xlab("Richness in both samples") +
  ylab("\nFold increase") + 
  labs(title = " ",
       subtitle = "Fungi") +
  coord_cartesian(ylim = c(1,3)) + 
  annotate(geom = "text", 
           x=750, 
           y=2.5, 
           label = paste0("r = ", 
                          round(cormod$estimate, 
                                digits = 2),
                          "\nP = ",
                          round(cormod$p.value,
                                digits = 2)), 
           fontface = "italic", size=3, color="black", 
           hjust = 0); fig3.c.f


```



```{r}
fig3 = ggarrange(fig3.a.p,fig3.b.p,fig3.c.p,
                 fig3.a.f,fig3.b.f,fig3.c.f,
                 nrow=2,ncol=3); fig3
```


















```{r}
# we're only interested in pots sampled twice in the coming model. pot 46 (p46) 
# lost its second sample during normalization, we therefore remove its first 
# sample from the dataset: 
p.rar = subset_samples(p.rar, pot_id!="p46"); p.rar # 

# make two phyloseq objects for the pots sampled twice - one for the first sample
# and one for the second sample. start with the first sample per pot:
dp = subset_samples(p.rar, interval=="double"); dp
dps = subset_samples(dp, sample_id!="1_2") # samples with sample_id = *_2 represent second sample per pot
dps = subset_samples(dps, sample_id!="2_2")
dps = subset_samples(dps, sample_id!="3_2")
dps = subset_samples(dps, sample_id!="4_2")
dps = subset_samples(dps, sample_id!="5_2") # filter out the second sample from every pot
dps = prune_taxa(taxa_sums(dps)>0, dps); dp; dps 

dps@sam_data$sampled_times = "one" # add a variable to distinguish from the second samples
dps@sam_data = sample_data(data.frame(cbind(dps@sam_data), 
                                      estimate_richness(dps, 
                                                        measures = "Observed"))); head(dps@sam_data) # estimate and record observed richness (ignore warning)

# create a data frame for easier modelling
obs.one.sams = data.frame(dps@sam_data) %>% 
  rename("obs_one_sample"="Observed"); head(obs.one.sams) # give Observed a better name


# now do the first+second sample:
dpm = merge_samples(dp, 
                    group = "pot_id"); dp; dpm # merge samples from corresponding pots

# fix pot_id and copy sample data from dps in a controlled manner:
dpm@sam_data$sampled_times = "two"; head(dpm@sam_data) # to distinguish from first samples
dpm@sam_data$pot_id = rownames(dpm@sam_data) 
dpm@sam_data$host = dps@sam_data$host
dpm@sam_data$container = dps@sam_data$container

dpm@sam_data = sample_data(data.frame(cbind(dpm@sam_data), 
                                      estimate_richness(dpm, 
                                                        measures = "Observed"))); head(dpm@sam_data) # estimate and record observed richness (ignore warning)

# create a data frame for easier modelling
obs.two.sams = data.frame(dpm@sam_data) %>% 
  rename("obs_two_samples"="Observed"); head(obs.two.sams) # give Observed a better name



# now combine the data into one data set:
fi.df = left_join(obs.one.sams, 
                  obs.two.sams, 
                  by = "pot_id") %>% 
  dplyr::select("pot_id", "host.x", "container.x", 
         "obs_one_sample", "obs_two_samples") %>% 
  dplyr::rename("container"="container.x",
                "host"="host.x") %>% 
  dplyr::mutate(fold_increase = obs_two_samples/obs_one_sample) 


# do some exploratory plotting:
ggarrange(
fi.df %>% 
  ggplot(aes(x=container, y=obs_one_sample)) + 
  geom_boxplot(),
fi.df %>% 
  ggplot(aes(x=container, y=obs_two_samples)) + 
  geom_boxplot(),
fi.df %>% 
  ggplot(aes(x=container, y=fold_increase)) + 
  geom_boxplot(),
nrow = 1)
#
ggarrange(
fi.df %>% 
  ggplot(aes(x=host, y=obs_one_sample)) + 
  geom_boxplot() + 
  facet_wrap(~container),
fi.df %>% 
  ggplot(aes(x=host, y=obs_two_samples)) + 
  geom_boxplot() + 
  facet_wrap(~container),
fi.df %>% 
  ggplot(aes(x=host, y=fold_increase)) + 
  geom_boxplot() + 
  facet_wrap(~container),
nrow = 3)
#


## fit a model: 

# fit the data to a suitable distribution: 
library(fitdistrplus) 

plotdist(fi.df$fold_increase, 
         histo = TRUE, demp = TRUE) # intial plotting

descdist(fi.df$fold_increase,
         boot = 1000) # skewness-kurtosis plot

x=fitdist(fi.df$fold_increase/10, 
          distr = "beta", method = "mle"); summary(x); plot(x) 

gofstat(x) # 

# fit a linear model to test the effect of container and host plant on the
# on observed species richness. include pot_id as a random effect to control
# for the fact that ca. half of the samples
library(lmerTest) # for fitting the model
library(effects) # for plotting effects of predictor variables


#

```



```{r}

# we're going to merge samples from the same pots. to do this, we first create a 
# new phyloseq object with only the samples we're going to merge: 
twice = subset_samples(fungi, interval == "double"); twice
twice.m = merge_samples(twice, group = "pot_id"); twice.m 

# reassign sample data in twice.m:
twice.m@sam_data$pot_id = rownames(twice.m@sam_data) # fix pot_id variable
twice.m@sam_data$pot_id = sub("p","",twice.m@sam_data$pot_id) 
twice.m@sam_data$pot_id = as.numeric(twice.m@sam_data$pot_id) # this will let us sort by pot_id
twice.m.sam = data.frame(twice.m@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id))

# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(sample_id!="1_2") %>%
  filter(sample_id!="2_2") %>%
  filter(sample_id!="3_2") %>%
  filter(sample_id!="4_2") %>%
  filter(sample_id!="5_2")

# and reassign variables: 
twice.m.sam$host = sam$host
twice.m.sam$container = sam$container
twice.m.sam$interval = sam$interval

# convert til sample_data() object:
twice.m.sam = sample_data(twice.m.sam); sample_names(twice.m.sam)

# and reassign to twice.m phyloseq object: 
sample_data(twice.m) = twice.m.sam; twice.m; head(twice.m@sam_data)


# now we want to get the merged samples back in the original dataset. we first
# create a new dataset with only samples from pots sampled once: 
once = subset_samples(fungi, interval == "single"); once

# and now we can merge the two phyloseq objects:
fungi2 = merge_phyloseq(twice.m, once); fungi2

sort(sample_sums(fungi2))
f2rar = rarefy_even_depth(fungi2, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE, 
                          sample.size = 237829); fungi2; f2rar

# now we're ready to have a look at observed species richness:
# make a dataframe for ease of testing and plotting:
obs.df = data.frame(cbind(estimate_richness(f2rar, 
                                            measures = "Observed"), 
                          sample_data(f2rar))); str(obs.df)

hist(obs.df$Observed); shapiro.test(obs.df$Observed) # not statistically significantly different from a normal distribution...
var.test(Observed ~ interval, obs.df) # variance is not significantly different between the two groups.
t.test(Observed ~ interval, obs.df, var.equal=TRUE)
si=obs.df %>% 
  filter(interval=="single"); sd(si$Observed); rm(si$Observed)
db=obs.df %>% 
  filter(interval=="double"); sd(db$Observed); rm(db$Observed)
#
obs.df %>% 
  ggplot(aes(x=interval, y=Observed, 
             fill=interval)) + 
  geom_boxplot(alpha = 0.6) + 
  scale_fill_manual(values = c("tan1", 
                               "gold1")) +
  theme_classic() + 
  ylab("Observed species richness") +
  scale_x_discrete(labels = c("Double", 
                              "Single")) +
  theme(legend.position = "none", 
        axis.title.x = element_blank())

```




###################################################

alpha diversity

PLANTS

we're interested in exploring the alpha diversity of contaminant plants and fungi.
we're going to fit models to determine which factors are important for determining
the diversity and composition of contaminant species in our samples. 

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(plants)); sort(sample_sums(plants))

# considering the read depth across all the samples, 8629 seems like a fair sample
# size to retain diversity (ASVs (species)) and power (samples) in the data set: 
p.rar = rarefy_even_depth(plants, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = 8629); plants; p.rar
```


model PLANT observed richness 

first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(p.rar,
                                         measures = "Observed"),
                       sample_data(p.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data, they're not significantly different from a normal
# distribution:
plotdist(log(obs$Observed), 
         histo = TRUE, demp = TRUE) # intial plotting

descdist(log(obs$Observed), 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed), 
          distr = "norm", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a normal distribution. 


# fit a linear model to test the effect of container and host plant on the
# on observed species richness. include pot_id as a random effect to control
# for the fact that ca. half of the samples
library(lmerTest) # for fitting the model
library(effects) # for plotting effects of predictor variables

p.mod1 = lmerTest::lmer(formula = log(Observed) ~ container + host + (1|pot_id),
                      data = obs); summary(p.mod1); plot(allEffects(p.mod1))

# we'll use DHARMa to validate the model's residuals:
library(DHARMa)
testDispersion(p.mod1)

sim.res.p.mod1 = simulateResiduals(p.mod1, 
                                   n = 250,
                                   plot = TRUE) # these simulated residuals look dandy

# review the model results:
p.mod1.res=summary(p.mod1); p.mod1.res
# seems host and host interacting with container had a significant on richness 
# in our samples. perform a post-hoc test with emmeans package:
library(emmeans)

p.posthoc = emmeans(p.mod1, 
                    list(pairwise ~ container*host), 
                    adjust = "Tukey"); p.posthoc # two significant pairwise comparisons

# make a nice figure to go with the results:
p = obs %>% 
  ggplot(aes(y=Observed, x=container,
             fill=host, group=interaction(container, 
                                          host))) +  
  theme_classic() +
  geom_boxplot(width=0.5, 
               outlier.shape = NA,
               aes(fill=host), 
               alpha=0.4) + 
  scale_fill_manual(values = c("chartreuse1",
                               "steelblue1")) + 
  scale_y_continuous(breaks = seq(0, 15, 3)) +
  theme(legend.title = element_blank(), 
        legend.position = c(0.85,0.85), 
        legend.background = element_rect(colour = "black",
                                         fill = "grey95"), 
        legend.text = element_text()) + 
  annotate(geom = "text", label = c("A", "B", "AB", "AB"),
           x = c(0.88, 1.12, 1.88, 2.12), 
           y = c(14.5, 7.5, 13.5, 9.5)); p
```



model PLANT fold increase from one to two soil samples per pot

we next want to find out if container and host predict the fold increase (FI) in
observed species richness when sampling pots first once and then twice. for this
model we will only include pots that were sampled twice, and we will fit mixed 
effects models with random slopes because the data are paired 
(observed richness in 1st sample vs observed richness in 1st + 2nd samples)

NEED TO NORMALIZE FIRST:
  - normalizing all sams prior to merging = taking and seqing 2 per pot = expensive <-- this one, right?
  - normalizing singles alone and merged alone prior to comparison = simulating osil volume increase without increasing sampling seqing effort = cheaper <-- but not exactly what we're testing here, right? should do BOTH

```{r}
# we're only interested in pots sampled twice in the coming model. pot 46 (p46) 
# lost its second sample during normalization, we therefore remove its first 
# sample from the dataset: 
p.rar = subset_samples(p.rar, pot_id!="p46"); p.rar # 

# make two phyloseq objects for the pots sampled twice - one for the first sample
# and one for the second sample. start with the first sample per pot:
dp = subset_samples(p.rar, interval=="double"); dp
dps = subset_samples(dp, sample_id!="1_2") # samples with sample_id = *_2 represent second sample per pot
dps = subset_samples(dps, sample_id!="2_2")
dps = subset_samples(dps, sample_id!="3_2")
dps = subset_samples(dps, sample_id!="4_2")
dps = subset_samples(dps, sample_id!="5_2") # filter out the second sample from every pot
dps = prune_taxa(taxa_sums(dps)>0, dps); dp; dps 

dps@sam_data$sampled_times = "one" # add a variable to distinguish from the second samples
dps@sam_data = sample_data(data.frame(cbind(dps@sam_data), 
                                      estimate_richness(dps, 
                                                        measures = "Observed"))); head(dps@sam_data) # estimate and record observed richness (ignore warning)

# create a data frame for easier modelling
obs.one.sams = data.frame(dps@sam_data) %>% 
  rename("obs_one_sample"="Observed"); head(obs.one.sams) # give Observed a better name


# now do the first+second sample:
dpm = merge_samples(dp, 
                    group = "pot_id"); dp; dpm # merge samples from corresponding pots

# fix pot_id and copy sample data from dps in a controlled manner:
dpm@sam_data$sampled_times = "two"; head(dpm@sam_data) # to distinguish from first samples
dpm@sam_data$pot_id = rownames(dpm@sam_data) 
dpm@sam_data$host = dps@sam_data$host
dpm@sam_data$container = dps@sam_data$container

dpm@sam_data = sample_data(data.frame(cbind(dpm@sam_data), 
                                      estimate_richness(dpm, 
                                                        measures = "Observed"))); head(dpm@sam_data) # estimate and record observed richness (ignore warning)

# create a data frame for easier modelling
obs.two.sams = data.frame(dpm@sam_data) %>% 
  rename("obs_two_samples"="Observed"); head(obs.two.sams) # give Observed a better name



# now combine the data into one data set:
fi.df = left_join(obs.one.sams, 
                  obs.two.sams, 
                  by = "pot_id") %>% 
  dplyr::select("pot_id", "host.x", "container.x", 
         "obs_one_sample", "obs_two_samples") %>% 
  dplyr::rename("container"="container.x",
                "host"="host.x") %>% 
  dplyr::mutate(fold_increase = obs_two_samples/obs_one_sample) 


# do some exploratory plotting:
ggarrange(
fi.df %>% 
  ggplot(aes(x=container, y=obs_one_sample)) + 
  geom_boxplot(),
fi.df %>% 
  ggplot(aes(x=container, y=obs_two_samples)) + 
  geom_boxplot(),
fi.df %>% 
  ggplot(aes(x=container, y=fold_increase)) + 
  geom_boxplot(),
nrow = 1)
#
ggarrange(
fi.df %>% 
  ggplot(aes(x=host, y=obs_one_sample)) + 
  geom_boxplot() + 
  facet_wrap(~container),
fi.df %>% 
  ggplot(aes(x=host, y=obs_two_samples)) + 
  geom_boxplot() + 
  facet_wrap(~container),
fi.df %>% 
  ggplot(aes(x=host, y=fold_increase)) + 
  geom_boxplot() + 
  facet_wrap(~container),
nrow = 3)
#


## fit a model: 

# fit the data to a suitable distribution: 
library(fitdistrplus) 

plotdist(fi.df$fold_increase, 
         histo = TRUE, demp = TRUE) # intial plotting

descdist(fi.df$fold_increase,
         boot = 1000) # skewness-kurtosis plot

x=fitdist(fi.df$fold_increase/10, 
          distr = "beta", method = "mle"); summary(x); plot(x) 

gofstat(x) # 

# fit a linear model to test the effect of container and host plant on the
# on observed species richness. include pot_id as a random effect to control
# for the fact that ca. half of the samples
library(lmerTest) # for fitting the model
library(effects) # for plotting effects of predictor variables


#

```




make extrapolated species accumulation curves

we're interested in approximating how many pots would have to be sampled per 
container to identify all contaminant species per container. we can extrapolate
based on our samples to get an estimate of this. we'll use the iNEXT package:

```{r}
library(iNEXT)

# we need to prepare our data for iNEXT. for doubts, refer to https://doi.org/10.17504/protocols.io.bu6fnzbn
# our data should be a species-by-samples matrix with incidence values, i.e. 
# 1 or 0 (presence or absence). species should be rows and samples be columns:

pnext = data.frame(t(plants@otu_table)) %>% 
  mutate_if(is.numeric, ~1 *(. !=0))
pnext = as.matrix(pnext) # 
pnext2 = list(pnext)

inext.out = iNEXT::iNEXT(pnext2, # 'Error in 1:ncol(x) : argument of length 0' when attempted to run with matrix
                         q = 0, 
                         datatype = "incidence_raw", 
                         knots = 200, 
                         se = TRUE 
                         )
ggiNEXT(inext.out, type = 1, se = TRUE)
```






FUNGI

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(fungi)); sort(sample_sums(fungi))

# we have many more reads per sample for fungi than for plants. we can normalize
# at the lowest sampling depth to retain all samples and almost most taxa. 
f.rar = rarefy_even_depth(fungi, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = min(sample_sums(fungi))); fungi; f.rar
```


model FUNGAL observed richness 


first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(f.rar,
                                         measures = "Observed"),
                       sample_data(f.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data and dividing by 10, they're not significantly different 
# from a beta distribution:
plotdist(log(obs$Observed)/10, 
         histo = TRUE, demp = TRUE) # when

descdist(log(obs$Observed)/10, 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed)/10, 
          distr = "beta", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a gamma distribution. 


# fit a simple model to test:
library(lmerTest) # used for the actual model
library(effects) # used for plotting effects of predictor variables
mod1 = lmerTest::lmer(formula = log(Observed) ~ container * host + (1|pot_id),
                      data = obs); summary(mod1); plot(allEffects(mod1))

# we'll also use DHARMa to validate the model:
library(DHARMa)
testDispersion(mod1)

simout = simulateResiduals(mod1, plot = F)
sort(residuals(simout))
plot(simout)
#

obs %>% 
ggplot(aes(y=Observed, x=host,
           fill=host)) + 
  geom_violin(alpha=0.5, width=0.7, 
              aes(fill=host)) + 
  geom_boxplot(width=0.2, outlier.shape = NA,
               aes(fill=NULL)) + 
  geom_jitter(width=0.4, pch=21, size=3, 
              aes(fill=host)) +
  scale_fill_manual(values = c("chartreuse1",
                               "chartreuse4")) + 
 # facet_wrap(~container) + 
  theme_classic() +
  theme(legend.position = "none")

```





FUNGI

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(fungi)); sort(sample_sums(fungi))

# we have many more reads per sample for fungi than for plants. we can normalize
# at the lowest sampling depth to retain all samples and almost most taxa. 
f.rar = rarefy_even_depth(fungi, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = min(sample_sums(fungi))); fungi; f.rar
```


model FUNGAL observed richness 


first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(f.rar,
                                         measures = "Observed"),
                       sample_data(f.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data and dividing by 10, they're not significantly different 
# from a beta distribution:
plotdist(log(obs$Observed)/10, 
         histo = TRUE, demp = TRUE) # when

descdist(log2(obs$Observed)/10, 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed)/10, 
          distr = "beta", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a gamma distribution. 


# fit a simple model to test:
library(glmmTMB)
library(effects)

mod1 = glmmTMB(data = obs, 
               formula = log(Observed)/10 ~ container * host + (1|pot_id),
               beta_family()); summary(mod1)


library(DHARMa)
testDispersion(mod1)

simout = simulateResiduals(mod1, plot = F)
sort(residuals(simout))
plot(simout)
#

obs %>% 
  ggplot(aes(y=Observed, x=container,
             fill=container)) +  
  theme_classic() +
  geom_boxplot(width=0.5, 
               outlier.shape = NA, 
               alpha=0.4) 
  scale_fill_manual(values = c("tan1",
                               "hotpink")) + 
#  scale_y_continuous(breaks = seq(0, 15, 3)) +
  theme(legend.title = element_blank(), 
        legend.position = c(0.85,0.85), 
        legend.background = element_rect(colour = "black",
                                         fill = "grey95"), 
        legend.text = element_text()) + 
  annotate(geom = "text", label = c("A", "B", "AB", "AB"),
           x = c(0.88, 1.12, 1.88, 2.12), 
           y = c(700, 550, 13.5, 9.5)); f

```




