---
title: "plant-import-paper-r-markdown"
author: "pgfarsund"
date: "2023-10-07"
output: github_document
---

```{r}
library(phyloseq)
library(tidyverse)
library(ggpubr)
library(effects)
library(BiodiversityR)
```


load, clean, and manipulate phyloseq objects:
```{r}
### plants

# load data:
plants = readRDS(file = "plants-phyloseq-object.RDS") # change the path to data's location on your computer

# manipulate the sample data:
sam = data.frame(plants@sam_data) %>% 
  select("sample","Location_name",                   # there is a lot of data we don't need
         "Specific_host",
         "Sample_ID") %>% 
  rename(c("container"="Location_name",              # rename columns for simplicity
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number
         sample=sub("_ITS_S2_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))); str(sam) 
sample_data(plants) = sample_data(sam); rm(sam); plants

# for this study we're only interested in vascular plants. our data set contains
# some bryophytes, green algae, additionally we're not interested in Taxus sp. or
# Thuja sp., because these were growing in the samples soil. 
# remove non-focal taxa:
plants=subset_taxa(plants, phylum=="Streptophyta")
plants=subset_taxa(plants, class!="Chlorophyceae")
plants=subset_taxa(plants, order!="Splachnales")
plants=subset_taxa(plants, order!="Bryales")
plants=subset_taxa(plants, order!="Dicranales")
plants=subset_taxa(plants, order!="Funariales")
plants=subset_taxa(plants, order!="Hypnales")
plants=subset_taxa(plants, order!="Klebsormidiaceae")
plants=subset_taxa(plants, order!="Pottiales")
plants=subset_taxa(plants, order!="Pseudoditrichales")
plants=subset_taxa(plants, species!="Thuja_sp.")
plants=subset_taxa(plants, species!="Thuja_occidentalis")
plants=subset_taxa(plants, species!="Taxus_sp.")
plants=subset_taxa(plants, species!="Taxus_canadensis"); plants



### fungi

# load data:
fungi = readRDS(file = "fungi-phyloseq-object.RDS") # change the path to data's location on your computer
fungi = subset_samples(fungi, negative_control=="FALSE")

# manipulate the sample data:
sam = data.frame(fungi@sam_data) %>% 
  select("sample","Location_name",                   # there is a lot of data we don't need
         "Specific_host",
         "Sample_ID") %>% 
  rename(c("container"="Location_name",              # rename columns for simplicity
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number
         sample=sub("_fITS7_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1)))
sample_data(fungi) = sample_data(sam); rm(sam); fungi
```





1. alpha diversity

PLANTS

we're interested in exploring the alpha diversity of contaminant plants and fungi.
we're going to fit models to determine which factors are important for determining
the diversity and composition of contaminant species in our samples. 

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(plants)); sort(sample_sums(plants))

# considering the read depth across all the samples, 8629 seems like a fair sample
# size to retain diversity (ASVs (species)) and power (samples) in the data set: 
p.rar = rarefy_even_depth(plants, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = 8629); plants; p.rar
```


model PLANT observed richness 

first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(p.rar,
                                         measures = "Observed"),
                       sample_data(p.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data, they're not significantly different from a normal
# distribution:
plotdist(log(obs$Observed), 
         histo = TRUE, demp = TRUE) # intial plotting

descdist(log(obs$Observed), 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed), 
          distr = "norm", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a normal distribution. 


# fit a simple model to test:
library(lmerTest) # used for the actual model
library(effects) # used for plotting effects of predictor variables
mod1 = lmerTest::lmer(formula = log(Observed) ~ container * host + (1|pot_id),
                      data = obs); summary(mod1); plot(allEffects(mod1))

# we'll also use DHARMa to validate the model:
library(DHARMa)
testDispersion(mod1)

simout = simulateResiduals(mod1, plot = F)
sort(residuals(simout))
plot(simout)
#

obs %>% 
ggplot(aes(y=Observed, x=host,
           fill=host)) + 
  geom_violin(alpha=0.5, width=0.7, 
              aes(fill=host)) + 
  geom_boxplot(width=0.2, outlier.shape = NA,
               aes(fill=NULL)) + 
  geom_jitter(width=0.4, pch=21, size=3, 
              aes(fill=host)) +
  scale_fill_manual(values = c("chartreuse1",
                               "chartreuse4")) + 
 # facet_wrap(~container) + 
  theme_classic() +
  theme(legend.position = "none")

```





FUNGI

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(fungi)); sort(sample_sums(fungi))

# we have many more reads per sample for fungi than for plants. we can normalize
# at the lowest sampling depth to retain all samples and almost most taxa. 
f.rar = rarefy_even_depth(fungi, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = min(sample_sums(fungi))); fungi; f.rar
```


model FUNGAL observed richness 


first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(f.rar,
                                         measures = "Observed"),
                       sample_data(f.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data and dividing by 10, they're not significantly different 
# from a beta distribution:
plotdist(log(obs$Observed)/10, 
         histo = TRUE, demp = TRUE) # when

descdist(log2(obs$Observed)/10, 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed)/10, 
          distr = "beta", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a gamma distribution. 


# fit a simple model to test:
library(glmmTMB)
library(effects)

mod1 = glmmTMB(data = obs, 
               formula = log(Observed)/10 ~ container * host + (1|pot_id),
               beta_family()); summary(mod1)


library(DHARMa)
testDispersion(mod1)

simout = simulateResiduals(mod1, plot = F)
sort(residuals(simout))
plot(simout)
#

obs %>% 
ggplot(aes(y=Observed, x=host)) + 
  geom_violin(alpha=0.5, width=0.7, 
              aes(fill=host)) + 
  geom_boxplot(width=0.2, outlier.shape = NA,
               aes(fill=NULL)) + 
  geom_jitter(width=0.4, pch=21, size=3, 
              aes(fill=host)) +
  scale_fill_manual(values = c("orange1",
                               "tan4")) + 
  facet_wrap(~container) + 
  theme_classic() +
  theme(legend.position = "none")

```




