---
title: "plant-import-paper-r-markdown"
author: "pgfarsund"
date: "2023"
output: github_document
---

this markdown document contains code for the statistical analyses reported in 
the paper preliminary titled "the plant import paper" (in prep for Biological 
Invasions). the markdown also contains code for analyses that are not implemented
in the manuscript at present, but might be included or at least included in the
supporting material. 

```{r}
library(phyloseq)
library(tidyverse)
library(ggpubr)
library(effects)
library(BiodiversityR)
```


load, clean, and manipulate phyloseq objects:
```{r}
### plants

# load data:
plants = readRDS(file = "plants-phyloseq-object.RDS") # change the path to data's location on your computer

# manipulate the sample data:
sam = data.frame(plants@sam_data) %>% 
  dplyr::select("sample","Location_name",            # there is a lot of data we don't need
                "Specific_host",
                "Sample_ID") %>% 
  rename(c("container"="Location_name",              # rename columns for simplicity
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number
         sample=sub("_ITS_S2_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))) %>%
  mutate(interval = c(rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10), 
                      rep("double", times=10),
                      rep("single", times=10))); str(sam) 
sample_data(plants) = sample_data(sam); rm(sam); plants

# for this study we're only interested in vascular plants. our data set contains
# some bryophytes, green algae, additionally we're not interested in Taxus sp. or
# Thuja sp., because these were growing in the samples soil. 
# remove non-focal taxa:
plants=subset_taxa(plants, phylum=="Streptophyta")
plants=subset_taxa(plants, class!="Chlorophyceae")
plants=subset_taxa(plants, order!="Splachnales")
plants=subset_taxa(plants, order!="Bryales")
plants=subset_taxa(plants, order!="Dicranales")
plants=subset_taxa(plants, order!="Funariales")
plants=subset_taxa(plants, order!="Hypnales")
plants=subset_taxa(plants, order!="Klebsormidiaceae")
plants=subset_taxa(plants, order!="Pottiales")
plants=subset_taxa(plants, order!="Pseudoditrichales")
plants=subset_taxa(plants, species!="Thuja_sp.")
plants=subset_taxa(plants, species!="Thuja_occidentalis")
plants=subset_taxa(plants, species!="Taxus_sp.")
plants=subset_taxa(plants, species!="Taxus_canadensis"); plants



### fungi

# load data:
fungi = readRDS(file = "fungi-phyloseq-object.RDS") # change the path to data's location on your computer
fungi = subset_samples(fungi, negative_control=="FALSE")

# manipulate the sample data:
sam = data.frame(fungi@sam_data) %>% 
  dplyr::select("sample","Location_name",            # there is a lot of data we don't need
                "Specific_host",
                "Sample_ID") %>%
  rename(c("container"="Location_name",              # rename columns for simplicity
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number
         sample=sub("_fITS7_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))) %>%
  mutate(interval = c(rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10), 
                      rep("double", times=10),
                      rep("single", times=10)))
sample_data(fungi) = sample_data(sam); rm(sam); fungi
```





1. alpha diversity

PLANTS

we're interested in exploring the alpha diversity of contaminant plants and fungi.
we're going to fit models to determine which factors are important for determining
the diversity and composition of contaminant species in our samples. 

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(plants)); sort(sample_sums(plants))

# considering the read depth across all the samples, 8629 seems like a fair sample
# size to retain diversity (ASVs (species)) and power (samples) in the data set: 
p.rar = rarefy_even_depth(plants, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = 8629); plants; p.rar
```


model PLANT observed richness 

first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(p.rar,
                                         measures = "Observed"),
                       sample_data(p.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data, they're not significantly different from a normal
# distribution:
plotdist(log(obs$Observed), 
         histo = TRUE, demp = TRUE) # intial plotting

descdist(log(obs$Observed), 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed), 
          distr = "norm", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a normal distribution. 


# fit a linear model to test the effect of container and host plant on the
# on observed species richness. include pot_id as a random effect to control
# for the fact that ca. half of the samples
library(lmerTest) # for fitting the model
library(effects) # for plotting effects of predictor variables

p.mod1 = lmerTest::lmer(formula = log(Observed) ~ container + host + (1|pot_id),
                      data = obs); summary(p.mod1); plot(allEffects(p.mod1))

# we'll use DHARMa to validate the model's residuals:
library(DHARMa)
testDispersion(p.mod1)

sim.res.p.mod1 = simulateResiduals(p.mod1, 
                                   n = 250,
                                   plot = TRUE) # these simulated residuals look dandy

# review the model results:
p.mod1.res=summary(p.mod1); p.mod1.res
# seems host and host interacting with container had a significant on richness 
# in our samples. perform a post-hoc test with emmeans package:
library(emmeans)

p.posthoc = emmeans(p.mod1, 
                    list(pairwise ~ container*host), 
                    adjust = "Tukey"); p.posthoc # two significant pairwise comparisons

# make a nice figure to go with the results:
p = obs %>% 
  ggplot(aes(y=Observed, x=container,
             fill=host, group=interaction(container, 
                                          host))) +  
  theme_classic() +
  geom_boxplot(width=0.5, 
               outlier.shape = NA,
               aes(fill=host), 
               alpha=0.4) + 
  scale_fill_manual(values = c("chartreuse1",
                               "steelblue1")) + 
  scale_y_continuous(breaks = seq(0, 15, 3)) +
  theme(legend.title = element_blank(), 
        legend.position = c(0.85,0.85), 
        legend.background = element_rect(colour = "black",
                                         fill = "grey95"), 
        legend.text = element_text()) + 
  annotate(geom = "text", label = c("A", "B", "AB", "AB"),
           x = c(0.88, 1.12, 1.88, 2.12), 
           y = c(14.5, 7.5, 13.5, 9.5)); p
```



model PLANT fold increase from one to two soil samples per pot

we next want to find out if container and host predict the fold increase (FI) in
observed species richness when sampling pots first once and then twice. for this
model we will only include pots that were sampled twice, and we will fit mixed 
effects models with random slopes because the data are paired 
(observed richness in 1st sample vs observed richness in 1st + 2nd samples)

NEED TO NORMALIZE FIRST:
  - normalizing all sams prior to merging = taking and seqing 2 per pot = expensive <-- this one, right?
  - normalizing singles alone and merged alone prior to comparison = simulating osil volume increase without increasing sampling seqing effort = cheaper <-- but not exactly what we're testing here, right? should do BOTH

```{r}
# we're only interested in pots sampled twice in the coming model. pot 46 (p46) 
# lost its second sample during normalization, we therefore remove its first 
# sample from the dataset: 
p.rar = subset_samples(p.rar, pot_id!="p46"); p.rar # 

# make two phyloseq objects for the pots sampled twice - one for the first sample
# and one for the second sample. start with the first sample per pot:
dp = subset_samples(p.rar, interval=="double"); dp
dps = subset_samples(dp, sample_id!="1_2") # samples with sample_id = *_2 represent second sample per pot
dps = subset_samples(dps, sample_id!="2_2")
dps = subset_samples(dps, sample_id!="3_2")
dps = subset_samples(dps, sample_id!="4_2")
dps = subset_samples(dps, sample_id!="5_2") # filter out the second sample from every pot
dps = prune_taxa(taxa_sums(dps)>0, dps); dp; dps 

dps@sam_data$sampled_times = "one" # add a variable to distinguish from the second samples
dps@sam_data = sample_data(data.frame(cbind(dps@sam_data), 
                                      estimate_richness(dps, 
                                                        measures = "Observed"))); head(dps@sam_data) # estimate and record observed richness (ignore warning)

# create a data frame for easier modelling
obs.one.sams = data.frame(dps@sam_data) %>% 
  rename("obs_one_sample"="Observed"); head(obs.one.sams) # give Observed a better name


# now do the first+second sample:
dpm = merge_samples(dp, 
                    group = "pot_id"); dp; dpm # merge samples from corresponding pots

# fix pot_id and copy sample data from dps in a controlled manner:
dpm@sam_data$sampled_times = "two"; head(dpm@sam_data) # to distinguish from first samples
dpm@sam_data$pot_id = rownames(dpm@sam_data) 
dpm@sam_data$host = dps@sam_data$host
dpm@sam_data$container = dps@sam_data$container

dpm@sam_data = sample_data(data.frame(cbind(dpm@sam_data), 
                                      estimate_richness(dpm, 
                                                        measures = "Observed"))); head(dpm@sam_data) # estimate and record observed richness (ignore warning)

# create a data frame for easier modelling
obs.two.sams = data.frame(dpm@sam_data) %>% 
  rename("obs_two_samples"="Observed"); head(obs.two.sams) # give Observed a better name



# now combine the data into one data set:
fi.df = left_join(obs.one.sams, 
                  obs.two.sams, 
                  by = "pot_id") %>% 
  dplyr::select("pot_id", "host.x", "container.x", 
         "obs_one_sample", "obs_two_samples") %>% 
  dplyr::rename("container"="container.x",
                "host"="host.x") %>% 
  dplyr::mutate(fold_increase = obs_two_samples/obs_one_sample) 


# do some exploratory plotting:
ggarrange(
fi.df %>% 
  ggplot(aes(x=container, y=obs_one_sample)) + 
  geom_boxplot(),
fi.df %>% 
  ggplot(aes(x=container, y=obs_two_samples)) + 
  geom_boxplot(),
fi.df %>% 
  ggplot(aes(x=container, y=fold_increase)) + 
  geom_boxplot(),
nrow = 1)
#
ggarrange(
fi.df %>% 
  ggplot(aes(x=host, y=obs_one_sample)) + 
  geom_boxplot() + 
  facet_wrap(~container),
fi.df %>% 
  ggplot(aes(x=host, y=obs_two_samples)) + 
  geom_boxplot() + 
  facet_wrap(~container),
fi.df %>% 
  ggplot(aes(x=host, y=fold_increase)) + 
  geom_boxplot() + 
  facet_wrap(~container),
nrow = 3)
#


## fit a model: 

# fit the data to a suitable distribution: 
library(fitdistrplus) 

plotdist(fi.df$fold_increase, 
         histo = TRUE, demp = TRUE) # intial plotting

descdist(fi.df$fold_increase,
         boot = 1000) # skewness-kurtosis plot

x=fitdist(fi.df$fold_increase, 
          distr = "", method = "mle"); summary(x); plot(x) 

gofstat(x) # 

# fit a linear model to test the effect of container and host plant on the
# on observed species richness. include pot_id as a random effect to control
# for the fact that ca. half of the samples
library(lmerTest) # for fitting the model
library(effects) # for plotting effects of predictor variables


#

```




make extrapolated species accumulation curves

we're interested in approximating how many pots would have to be sampled per 
container to identify all contaminant species per container. we can extrapolate
based on our samples to get an estimate of this. we'll use the iNEXT package:

```{r}
library(iNEXT)

# we need to prepare our data for iNEXT. for doubts, refer to https://doi.org/10.17504/protocols.io.bu6fnzbn
# our data should be a species-by-samples matrix with incidence values, i.e. 
# 1 or 0 (presence or absence). species should be rows and samples be columns:

pnext = data.frame(t(plants@otu_table)) %>% 
  mutate_if(is.numeric, ~1 *(. !=0))
pnext = as.matrix(pnext) # 
pnext2 = list(pnext)

inext.out = iNEXT::iNEXT(pnext2, # 'Error in 1:ncol(x) : argument of length 0' when attempted to run with matrix
                         q = 0, 
                         datatype = "incidence_raw", 
                         knots = 200, 
                         se = TRUE 
                         )
ggiNEXT(inext.out, type = 1, se = TRUE)
```






FUNGI

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(fungi)); sort(sample_sums(fungi))

# we have many more reads per sample for fungi than for plants. we can normalize
# at the lowest sampling depth to retain all samples and almost most taxa. 
f.rar = rarefy_even_depth(fungi, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = min(sample_sums(fungi))); fungi; f.rar
```


model FUNGAL observed richness 


first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(f.rar,
                                         measures = "Observed"),
                       sample_data(f.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data and dividing by 10, they're not significantly different 
# from a beta distribution:
plotdist(log(obs$Observed)/10, 
         histo = TRUE, demp = TRUE) # when

descdist(log2(obs$Observed)/10, 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed)/10, 
          distr = "beta", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a gamma distribution. 


# fit a simple model to test:
library(lmerTest) # used for the actual model
library(effects) # used for plotting effects of predictor variables
mod1 = lmerTest::lmer(formula = log(Observed) ~ container * host + (1|pot_id),
                      data = obs); summary(mod1); plot(allEffects(mod1))

# we'll also use DHARMa to validate the model:
library(DHARMa)
testDispersion(mod1)

simout = simulateResiduals(mod1, plot = F)
sort(residuals(simout))
plot(simout)
#

obs %>% 
ggplot(aes(y=Observed, x=host,
           fill=host)) + 
  geom_violin(alpha=0.5, width=0.7, 
              aes(fill=host)) + 
  geom_boxplot(width=0.2, outlier.shape = NA,
               aes(fill=NULL)) + 
  geom_jitter(width=0.4, pch=21, size=3, 
              aes(fill=host)) +
  scale_fill_manual(values = c("chartreuse1",
                               "chartreuse4")) + 
 # facet_wrap(~container) + 
  theme_classic() +
  theme(legend.position = "none")

```





FUNGI

first normalize the number of sequences across samples:

```{r}
quantile(sample_sums(fungi)); sort(sample_sums(fungi))

# we have many more reads per sample for fungi than for plants. we can normalize
# at the lowest sampling depth to retain all samples and almost most taxa. 
f.rar = rarefy_even_depth(fungi, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE,
                          sample.size = min(sample_sums(fungi))); fungi; f.rar
```


model FUNGAL observed richness 


first we want to know which factors are important for the alpha diversity in each
individual soil sample. we therefore include all samples from all pots, and will
include pot_id as a random effect in our model to account for the fact that

```{r}
# create a data.frame with observed richness and sampling data:
obs = cbind(data.frame(estimate_richness(f.rar,
                                         measures = "Observed"),
                       sample_data(f.rar))); str(obs)

# we do some plotting to explore the data:
obs %>% 
  ggplot(aes(x=container, y=Observed, 
             fill=host)) + 
  geom_boxplot()

# it seems like there is a difference in observed species richness between thuja
# and taxus in container 1 but not in container 2. this indicates an interaction 
# effect between container and host. 

# before fitting our data to a model, we need to figure out what distribution
# bets fits our data. we use the package fitdistrplus for this:
library(fitdistrplus) # load package

# we found it difficult to fit the data to a suitable distribution. by log
# transforming the data and dividing by 10, they're not significantly different 
# from a beta distribution:
plotdist(log(obs$Observed)/10, 
         histo = TRUE, demp = TRUE) # when

descdist(log2(obs$Observed)/10, 
         boot = 1000) # skewness-kurtosis plot

x=fitdist(log(obs$Observed)/10, 
          distr = "beta", 
          method = "mle"); summary(x); plot(x) 

gofstat(x) # not statistically significantly different from a gamma distribution. 


# fit a simple model to test:
library(glmmTMB)
library(effects)

mod1 = glmmTMB(data = obs, 
               formula = log(Observed)/10 ~ container * host + (1|pot_id),
               beta_family()); summary(mod1)


library(DHARMa)
testDispersion(mod1)

simout = simulateResiduals(mod1, plot = F)
sort(residuals(simout))
plot(simout)
#

obs %>% 
ggplot(aes(y=Observed, x=host)) + 
  geom_violin(alpha=0.5, width=0.7, 
              aes(fill=host)) + 
  geom_boxplot(width=0.2, outlier.shape = NA,
               aes(fill=NULL)) + 
  geom_jitter(width=0.4, pch=21, size=3, 
              aes(fill=host)) +
  scale_fill_manual(values = c("orange1",
                               "tan4")) + 
  facet_wrap(~container) + 
  theme_classic() +
  theme(legend.position = "none")

```




