---
title: "plant-import-paper-r-markdown"
author: "pgfarsund"
date: "2023"
output: github_document
---

this markdown document contains code for the analysis and results already implemented 
in the manuscript for "the plant import paper" (in prep for Biological Invasions). in 
addition the these results, another script ("additional-analyses-for-the-plant-import-paper") 
contains analyses of the same data that have not (yet) been implemented in the manuscript.

```{r}
library(phyloseq)
library(tidyverse)
library(ggpubr)
```


load and manipulate phyloseq objects:
```{r}
### fungi

# load data:
plants = readRDS(file = "plants-phyloseq-object.RDS") # change the path to data's location on your computer.

# manipulate the sample data:
sam = data.frame(plants@sam_data) %>% 
  dplyr::select("sample","Location_name",            # there is a lot of data we don't need.
                "Specific_host",
                "Sample_ID") %>% 
  rename(c("container"="Location_name",              # rename columns for simplicity.
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number.
         sample=sub("_ITS_S2_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID.
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot.
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))) %>%
  mutate(interval = c(rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10), 
                      rep("double", times=10),
                      rep("single", times=10))); str(sam) 
sample_data(plants) = sample_data(sam); rm(sam); plants

# for this study we're only interested in vascular plants. our data set contains
# some bryophytes and green algae, additionally we're not interested in Taxus sp. 
# or Thuja sp., because these were ornamentals in the sampled soil. 
# remove non-focal taxa:
plants=subset_taxa(plants, phylum=="Streptophyta")
plants=subset_taxa(plants, class!="Chlorophyceae")
plants=subset_taxa(plants, order!="Splachnales")
plants=subset_taxa(plants, order!="Bryales")
plants=subset_taxa(plants, order!="Dicranales")
plants=subset_taxa(plants, order!="Funariales")
plants=subset_taxa(plants, order!="Hypnales")
plants=subset_taxa(plants, order!="Klebsormidiaceae")
plants=subset_taxa(plants, order!="Pottiales")
plants=subset_taxa(plants, order!="Pseudoditrichales")
plants=subset_taxa(plants, species!="Thuja_sp.")
plants=subset_taxa(plants, species!="Thuja_occidentalis")
plants=subset_taxa(plants, species!="Taxus_sp.")
plants=subset_taxa(plants, species!="Taxus_canadensis"); plants



### fungi

# load data:
fungi = readRDS(file = "fungi-phyloseq-object.RDS") # change the path to data's location on your computer
fungi = subset_samples(fungi, negative_control=="FALSE")

# manipulate the sample data:
sam = data.frame(fungi@sam_data) %>% 
  dplyr::select("sample","Location_name",            # there is a lot of data we don't need
                "Specific_host",
                "Sample_ID") %>%
  rename(c("container"="Location_name",              # rename columns for simplicity
           "host"="Specific_host", 
           "sample_id"="Sample_ID")) %>% 
  mutate(sample=sub("PLANTEIMP_21_", "", sample),    # easier to interpret sample number
         sample=sub("_fITS7_ITS4", "", sample)) %>% 
  mutate(container = sub("Bil ", "", container)) %>% # remove "Bil " from the container ID
  mutate(pot_id = c(rep(paste0("p", 1:5), each=2),   # add an ID to each pot - this will be
                    rep(paste0("p", 6:15), each=1),  # identical between samples taken from
                    rep(paste0("p", 16:20), each=2), # the same pot
                    rep(paste0("p", 21:30), each=1),
                    rep(paste0("p", 31:35), each=2),
                    rep(paste0("p", 36:45), each=1),
                    rep(paste0("p", 46:50), each=2),
                    rep(paste0("p", 51:60), each=1))) %>%
  mutate(interval = c(rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10),
                      rep("double", times=10),
                      rep("single", times=10), 
                      rep("double", times=10),
                      rep("single", times=10)))
sample_data(fungi) = sample_data(sam); rm(sam); fungi
```


##### RESULTS

### contaminant plants and fungi ### (Fig. 1)

we identified many contaminant plants and fungi that were native, known aliens, 
and potential alien species to Norway. we've grouped these species as "Native", 
"Alien", "Swe., Fin., Den." (i.e. native to Norway's neighboring countries), 
(native to) "Europe", and (native to) "Non-Europe". we will make a stacked bar
chart to show the relative frequencies of these different categories:

```{r}

# build the data frame:
df = data.frame("freq" = c(30.30, 45.45, 0, 12.12, 12.12, 0, # start with plant frequencies
                           39.53, 0.81, 22.32, 22.89, 10.88, 3.57), # then fungal frequencies
                "group" = rep(c("Plants",
                                "Fungi"), each=6),
                "cats" = c(rep(c("Native",
                                 "Alien",
                                 "Swe., Fin., Den.",
                                 "Europe", 
                                 "Non-Europe",
                                 "NA"), times=2)))

# make Figure 1 in the manuscript:
fig.1 = df %>%
ggplot(aes(x=factor(group, 
                    level=c("Plants",
                            "Fungi")), # order the groups on the x-axis
           y=freq, 
           fill=factor(cats, 
                       levels = c("Native", # we want the stacked bars to have this specific order
                                  "Alien", 
                                  "Swe., Fin., Den.",
                                  "Europe", 
                                  "Non-Europe", 
                                  "NA")),
           color=factor(cats, 
                        levels = c("Native", 
                                   "Alien", 
                                   "Swe., Fin., Den.",
                                   "Europe", 
                                   "Non-Europe", 
                                   "NA")))) + 
  geom_col(position = "fill", 
           width = 0.55) + 
  theme_classic(base_size = 15) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", # some nice colors to fill the bars
                               "#F0E442", "#CC79A7", "#999999")) +
  scale_color_manual(values = rep("black", times = 6)) +
  theme(legend.position = "right") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.title = element_blank()) +
  annotate(geom = "text", 
           label = c("30.3 %", 
                     "45.5 %",
                     "12.1 %",
                     "12.1 %", 
                     "39.5 %",
                     "0.8 %",
                     "22.3 %", 
                     "22.9 %", 
                     "10.9 %",
                     "3.6 %"),
           x = rep(c(1,2,2.45,2,2.45), # we want "0.8 %" and "3.6 %" outside the actual bar, which makes this a bit tricky
                   times = c(4,1,1,3,1)), 
           y = c(0.84, 0.45, 0.18, 0.06, # plant percent labels
                 0.8, 0.6, 0.5, 0.26, 0.1, 0.02), # fungi percent labels
           size = 4, color = "black", fontface = 1) + 
  annotate(geom = "segment", 
           x = rep(2.25,
                   times=2), 
           xend = rep(2.315, 
                      times=2), 
           y = c(0.60,
                 0.02), 
           yend = c(0.60,
                    0.02)) + 
  scale_y_continuous(labels = scales::percent); fig.1 # save as .pdf landscape 5 x 6

```


### metabarcoding vs. germination ### (Fig. 2)

we identified 33 plant species with metabarcoding and 30 with germination. five 
species were identified by both methods, adding up to 58 contaminant plant species
identified all together. we also checked whether identified plants were registered 
as products in the online assortments of Noviflora and Plantasjen in the fall of 
2021 (the year of sampling). we will make two stacked bar charts to show how many 
species fall into each category, while also showing how many species were identified
with metabarcoding, germination, or both: 

```{r}
df2 = read.csv(file = "C:/Users/pgfar/Dropbox/Whats_in_the_pot/plant-import-paper/fig_2_dataset.csv",
               sep = ";"); str(df2) # change path 

my_colors = c("#CC79A7",
              "#009E73",
              "#E69F00")

a = df2 %>% 
  mutate(prev_det=sub("TRUE", "Yes", prev_det)) %>% 
  mutate(prev_det=sub("FALSE", "No", prev_det)) %>% 
  ggplot(aes(x=factor(prev_det, 
                             levels = c("Yes", 
                                        "No")),
                color=factor(id_method,
                             levels = c("Metabarcoding",
                                        "Germination",
                                        "Both")),
                fill=factor(id_method,
                            levels = c("Metabarcoding",
                                       "Germination",
                                       "Both")))) + 
  geom_bar(position = "stack",
           width = 0.55) +
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = rep("black", times=3)) +
  theme_classic(base_size = 15) + 
  xlab("Previously identified?") + 
  ylab("Number of plant species") + 
  labs(fill = "Identification\nmethod", 
       color = "Identification\nmethod", 
       subtitle = "a.") +
  theme(legend.title = element_text(size = 13)) +
  scale_y_continuous(breaks = seq(0, 45, 10), 
                     limits = c(0, 45)); a

b = df2 %>% 
  mutate(ornamental=sub("TRUE", "Yes", ornamental)) %>% 
  mutate(ornamental=sub("FALSE", "No", ornamental)) %>% 
  ggplot(aes(x=factor(ornamental, 
                             levels = c("Yes", 
                                        "No")),
                color=factor(id_method,
                           levels = c("Metabarcoding",
                                      "Germination",
                                      "Both")),
                fill=factor(id_method,
                           levels = c("Metabarcoding",
                                      "Germination",
                                      "Both")))) + 
  geom_bar(position = "stack",
           width = 0.55) +
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = rep("black", times=3)) +
  theme_classic(base_size = 15) + 
  xlab("Ornamental?") + 
  ylab(" ") + 
  labs(fill = "Identification\nmethod", 
       color = "Identification\nmethod", 
       subtitle = "b.") +
  theme(legend.title = element_text(size = 13)) +
  scale_y_continuous(breaks = seq(0, 55, 10), 
                     limits = c(0, 45)); b

fig.2 = ggarrange(a, b, 
                    common.legend = TRUE,
                    legend = "right", 
                    nrow = 1); fig.2

```


### effect of sampling strategy on contaminant species richness ### (Fig. 3)

we want to investigate the effect of sampling effort on the observed species 
richness of contaminant plants and fungi identified by metabarcoding. we'll do
this in two steps:

1. we want to know whether a larger volume of soil per sample had an effect on 
the observed species richness per pot. unfortunately, all samples consited of 
approximately 25 grams. we're therefore going to simulate doubling the soil sample 
volume by merging samples that were taken from the same same pot, prior to 
normalizing the read counts across all samples:

```{r}

# we're going to merge samples from the same pots. to do this, we first create a 
# new phyloseq object with only the samples we're going to merge: 
twice = subset_samples(plants, interval == "double"); twice
twice.m = merge_samples(twice, group = "pot_id"); twice.m 

# reassign sample data in twice.m:
twice.m@sam_data$pot_id = rownames(twice.m@sam_data) # fix pot_id variable
twice.m@sam_data$pot_id = sub("p","",twice.m@sam_data$pot_id) 
twice.m@sam_data$pot_id = as.numeric(twice.m@sam_data$pot_id) # this will let us sort by pot_id
twice.m.sam = data.frame(twice.m@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id))

# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(sample_id!="1_2") %>%
  filter(sample_id!="2_2") %>%
  filter(sample_id!="3_2") %>%
  filter(sample_id!="4_2") %>%
  filter(sample_id!="5_2")

# and reassign variables: 
twice.m.sam$host = sam$host
twice.m.sam$container = sam$container
twice.m.sam$interval = sam$interval

# convert til sample_data() object:
twice.m.sam = sample_data(twice.m.sam); sample_names(twice.m.sam)

# and reassign to twice.m phyloseq object: 
sample_data(twice.m) = twice.m.sam; twice.m; head(twice.m@sam_data)


# now we want to get the merged samples back in the original dataset. we first
# create a new dataset with only samples from pots sampled once: 
once = subset_samples(plants, interval == "single"); once

# and now we can merge the two phyloseq objects:
plants2 = merge_phyloseq(twice.m, once); plants2

sort(sample_sums(plants2))
f2rar = rarefy_even_depth(plants2, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE, 
                          sample.size = 14215); plants2; f2rar

# now we're ready to have a look at observed species richness:
# make a dataframe for ease of testing and plotting:
obs.df = data.frame(cbind(estimate_richness(f2rar, 
                                            measures = "Observed"), 
                          sample_data(f2rar))); str(obs.df)

hist(obs.df$Observed); shapiro.test(obs.df$Observed) # not statistically significantly different from a normal distribution...
var.test(Observed ~ interval, obs.df) # the variance in the two groups is not significantly different from each other
mod1=t.test(Observed ~ interval, obs.df, var.equal=TRUE)
si=obs.df %>% 
  filter(interval=="single"); sd(si$Observed)
db=obs.df %>% 
  filter(interval=="double"); sd(db$Observed)

# make a nice boxplot to go with the results:
fig3.a.p=obs.df %>% 
  ggplot(aes(x=interval, y=Observed, 
             fill=interval)) + 
  geom_boxplot(alpha = 0.6) + 
  scale_fill_manual(values = c("chartreuse2", 
                               "steelblue1")) +
  theme_classic() + 
  ylab("Observed species richness") +
  scale_x_discrete(labels = c("Double", 
                              "Single")) +
  theme(legend.position = "none") +
  labs(title = "A.", 
       subtitle = "Plants") +
  xlab("Soil sample volume") +
  annotate(geom = "text", 
           x=1.5, 
           y=15, 
           label = paste0("P = ",
                          round(mod1$p.value, 
                                digits = 3)), # annotate rho and p-value
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0); fig3.a.p

```

to go along with the results about pots sampled twice, we're interested in the
overall recapture rate of the species that were identified in pots sampled twice.
i.e., were species identified consistently in both samples from the pots they
were identified in? or rather, was two samples (or increased volume) necessary
for identifying certain species in many pots?

```{r}
# we will make a data frame of species presence/absence in first and second samples
# from every pot, making it easy for us to see whether different species were 
# identified in none, one, or both, samples from every pot. 

twice = subset_samples(plants, interval == "double"); twice # subset pots sampled twice

first = subset_samples(twice, grepl(pattern = "_1", # subset first samples
                                    x = twice@sam_data$sample_id)); first = prune_taxa(taxa_sums(first)>0,
                                                                                       first); first
first@sam_data$turn = "first"; head(first@sam_data) # assign variable to distinguish from "second" samples
first@sam_data$row = paste0(first@sam_data$pot_id, # pot ID
                            "_",                   # _
                            first@sam_data$turn)   # turn

second = subset_samples(twice, grepl(pattern = "_2", # second samples
                                    x = twice@sam_data$sample_id)); second = prune_taxa(taxa_sums(second)>0,
                                                                                        second); second
second@sam_data$turn = "second"; head(second@sam_data)
second@sam_data$row = paste0(second@sam_data$pot_id, # pot ID
                             "_",                    # _
                             second@sam_data$turn)   # turn

twice = merge_phyloseq(first, 
                       second); twice # merge back together

sample_names(twice) = twice@sam_data$row; head(twice@sam_data) # assign the new sample names

colnames(twice@otu_table) = twice@tax_table[,6]

# now we make the new data frame with presence/absence data:
pres.abs.df = data.frame(twice@otu_table) %>% 
  mutate_if(is.numeric, ~1 * (. > 0)) 

pres.abs.df.t = data.frame(t(pres.abs.df)) # transpose to make pots columns

# split first and second samples and clean up the colnames before summarizing:
first = data.frame(pres.abs.df.t) %>% 
  select(matches("_first"))
colnames(first) = sub("_first", "", colnames(first))

second = data.frame(pres.abs.df.t) %>% 
  select(matches("_second"))
colnames(second) = sub("_second", "", colnames(second))

# by summarizing the two data frames, we'll be able to see if species were identified
# 0, 1, or 2 times per pot:
pres.abs.df.fin = data.frame(first+second)

# this is where you find the overall recapture rate of 70.3 % 
table(pres.abs.df.fin==0) # 175 identifications in total (i.e. != 0)
table(pres.abs.df.fin==1) # species X was identified in only one sample per pot a total of 123 times
table(pres.abs.df.fin==2) # species X was identified in both samples per pot a total of 52 times

# check recapture rate for two known alien species to Norway:
con = pres.abs.df.fin[rownames(pres.abs.df.fin) %in% "Erigeron_canadensis", ] # E. canadensis is synonymous to Conyza canadensis
table(con==0) 
table(con==1) 
table(con==2)

acer = pres.abs.df.fin[rownames(pres.abs.df.fin) %in% "Acer_pseudoplatanus", ]
acer = pres.abs.df.fin["Acer_pseudoplatanus",]
table(acer==0) 
table(acer==1) 
table(acer==2)

```

2. next we want to calculate the fold increase in species richness per pot when
increasing from one to two samples. to do this we will only use data from pots
sampled twice. FI = fold increase, S = sample, Obs = observed species richness:
FI = (Obs S1 + Obs S2) / Obs S1

```{r}
twice = subset_samples(plants, interval == "double"); twice = prune_taxa(taxa_sums(twice)>0,
                                                                         twice); twice
first = subset_samples(twice, grepl(pattern = "_1", 
                                    x = twice@sam_data$sample_id)); first = prune_taxa(taxa_sums(first)>0,
                                                                                       first); first
first@sam_data$turn = "first"; head(first@sam_data) # add a variable to distinguish between first and merged samples
second = merge_samples(twice, group = "pot_id"); second 

# reassign sample data in second:
second@sam_data$pot_id = rownames(second@sam_data) # fix pot_id variable
second@sam_data$pot_id = sub("p","", second@sam_data$pot_id) 
second@sam_data$pot_id = as.numeric(second@sam_data$pot_id) # this will let us sort by pot_id
second.sam = data.frame(second@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id)); head(second.sam); tail(second.sam)



# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(grepl(pattern = "_2", 
               sam$sample_id))

# and reassign variables: 
second.sam$host = sam$host
second.sam$container = sam$container
second.sam$interval = sam$interval
second.sam$turn = "second"; # add a variable to distinguish between merged and first samples

# convert til sample_data() object:
second.sam = sample_data(second.sam); sample_names(second.sam)

# and reassign to twice.m phyloseq object: 
sample_data(second) = second.sam; twice.m; head(second@sam_data)


# now we can merge the two phyloseq objects in order to normalize them:
fiphy = merge_phyloseq(first, second); fiphy

# normalize the new phyloseq object:
sort(sample_sums(fiphy)) # we can normalize at the lowest read depth to keep all samples and only lose two taxa:
fiphyrar = rarefy_even_depth(fiphy, 
                             rngseed = 123, 
                             replace = FALSE, 
                             trimOTUs = TRUE, 
                             sample.size = 10378); fiphy; fiphyrar

# now we have to separate the new object into two dataframes before we combine
# them to calculate and plot fold increase: 
first = subset_samples(fiphyrar, turn=="first")
first.df = data.frame(cbind(estimate_richness(first, 
                                              measures = "Observed"),
                            first@sam_data)) %>% 
  rename("obs_first"="Observed",
         "pot_id_first"="pot_id",
         "turn_first"="turn") %>% 
  mutate(pot_id=pot_id_first); head(first.df)

second = subset_samples(fiphyrar, turn=="second")
second.df = data.frame(cbind(estimate_richness(second, 
                                              measures = "Observed"),
                            second@sam_data)) %>% 
  rename("obs_second"="Observed",
         "pot_id_second"="pot_id",
         "turn_second"="turn") %>% 
  mutate(pot_id=pot_id_second); head(second.df)

fidf = full_join(first.df, second.df, by = "pot_id") %>% 
  select("obs_first", "obs_second", 
         "pot_id", "pot_id_first", "pot_id_second", 
         "turn_first", "turn_second")

fidf = fidf %>% 
  mutate(foldincrease = obs_second/obs_first)
fig3.b.p=ggplot(fidf, aes(x=reorder(pot_id, -foldincrease), 
                 y=foldincrease,
                 fill="darkolivegreen1", color="black")) + 
  geom_col(width=1) +
  scale_fill_manual(values = "darkolivegreen1") +
  scale_color_manual(values = "black") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.ticks.x = element_blank()
      ) + 
  ylab("Fold increase") +
  xlab("Sampled pots") +
  scale_x_discrete(labels = paste0(rep(" ", times=20))) +
  labs(title = "B.",
       subtitle = "Plants") +
  annotate(geom = "text", 
           x=5, 
           y=2.5, 
           label = paste0("Median = ", 
                         round(median(fidf$foldincrease), 
                               digits = 2)), # calculate and annotate manually...
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0) +
  scale_y_continuous(breaks = seq(0, 3, 0.5), limits = c(0, 3)) + 
  coord_cartesian(ylim = c(0.9, 3)); fig3.b.p

# additionally, we will investigate whether there is a correlation between 
# the richness in the second sample and fold increase for each pot:
cormod = cor.test(fidf$foldincrease, fidf$obs_second, 
         method = "pearson")
fig3.c.p=ggplot(fidf, aes(x=obs_second, 
                 y=foldincrease,
                 fill="darkolivegreen1")) + 
  theme_classic() +
  theme(legend.position = "none", 
        ) +
  scale_color_manual(values = "black") + 
  scale_fill_manual(values = "darkolivegreen1") +
  geom_point(size=6, pch=21) + 
  geom_smooth(method = "lm", 
              color="black", 
              fill="grey90") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 4, 0.5), 
                    limits = c(0, 3.6)) +
  xlab("Richness in both samples") +
  ylab("\nFold increase") + 
  labs(title = "C.",
       subtitle = "Plants") +
  coord_cartesian(ylim = c(1,3)) + 
  annotate(geom = "text", 
           x=8.8, 
           y=2.8, 
           label = paste0("r = ", 
                          round(cormod$estimate, 
                                digits = 3),
                          "\nP = ",
                          round(cormod$p.value,
                                digits = 3)), 
           fontface = "italic", size=3, color="black", 
           hjust = 0); fig3.c.p


```

we now run the three previous chunks for fungi:

1. soil volume:

```{r}

# we're going to merge samples from the same pots. to do this, we first create a 
# new phyloseq object with only the samples we're going to merge: 
twice = subset_samples(fungi, interval == "double"); twice
twice.m = merge_samples(twice, group = "pot_id"); twice.m 

# reassign sample data in twice.m:
twice.m@sam_data$pot_id = rownames(twice.m@sam_data) # fix pot_id variable
twice.m@sam_data$pot_id = sub("p","",twice.m@sam_data$pot_id) 
twice.m@sam_data$pot_id = as.numeric(twice.m@sam_data$pot_id) # this will let us sort by pot_id
twice.m.sam = data.frame(twice.m@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id))

# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(sample_id!="1_2") %>%
  filter(sample_id!="2_2") %>%
  filter(sample_id!="3_2") %>%
  filter(sample_id!="4_2") %>%
  filter(sample_id!="5_2")

# and reassign variables: 
twice.m.sam$host = sam$host
twice.m.sam$container = sam$container
twice.m.sam$interval = sam$interval

# convert til sample_data() object:
twice.m.sam = sample_data(twice.m.sam); sample_names(twice.m.sam)

# and reassign to twice.m phyloseq object: 
sample_data(twice.m) = twice.m.sam; twice.m; head(twice.m@sam_data)


# now we want to get the merged samples back in the original dataset. we first
# create a new dataset with only samples from pots sampled once: 
once = subset_samples(fungi, interval == "single"); once

# and now we can merge the two phyloseq objects:
fungi2 = merge_phyloseq(twice.m, once); fungi2

sort(sample_sums(fungi2))
f2rar = rarefy_even_depth(fungi2, 
                          rngseed = 123, 
                          replace = FALSE, 
                          trimOTUs = TRUE, 
                          sample.size = 237829); fungi2; f2rar

# now we're ready to have a look at observed species richness:
# make a dataframe for ease of testing and plotting:
obs.df = data.frame(cbind(estimate_richness(f2rar, 
                                            measures = "Observed"), 
                          sample_data(f2rar))); str(obs.df)

hist(obs.df$Observed); shapiro.test(obs.df$Observed) # not statistically significantly different from a normal distribution...
var.test(Observed ~ interval, obs.df) # the variance in the two groups is not significantly different from each other
mod1=t.test(Observed ~ interval, 
            obs.df, var.equal=TRUE)
si=obs.df %>% 
  filter(interval=="single"); sd(si$Observed)
db=obs.df %>% 
  filter(interval=="double"); sd(db$Observed)

# make a nice boxplot to go with the results:
fig3.a.f=obs.df %>% 
  ggplot(aes(x=interval, y=Observed, 
             fill=interval)) + 
  geom_boxplot(alpha = 0.6) + 
  scale_fill_manual(values = c("tan1", 
                               "gold1")) +
  theme_classic() + 
  ylab("Observed species richness") +
  scale_x_discrete(labels = c("Double", 
                              "Single")) +
  theme(legend.position = "none") +
  labs(title = " ", 
       subtitle = "Fungi") +
  xlab("Soil sample volume") +
  annotate(geom = "text", 
           x=1.5, 
           y=850, 
           label = paste0("P = <",
                          round(mod1$p.value, 
                                digits = 3),
                          ".001"), # annotate rho and p-value
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0); fig3.a.f

```

overall recapture rate:
!!!THIS CHUNK NEEDS CLEANING!!!
```{r}

#!!!THIS CHUNK NEEDS CLEANING!!!
#!!!THIS CHUNK NEEDS CLEANING!!!
#!!!THIS CHUNK NEEDS CLEANING!!!

x = df

pf = subset_samples(x, turn=="first")

pf@sam_data$row = paste0(pf@sam_data$pot_id,
                        "_", 
                        pf@sam_data$turn); head(pf@sam_data)

psec = subset_samples(x, turn=="second")
psec@sam_data$row = paste0(psec@sam_data$pot_id,
                           "_", 
                           psec@sam_data$turn); head(psec@sam_data)

x = merge_phyloseq(pf, psec); x
sample_names(x) = x@sam_data$row

colnames(x@otu_table) = data.frame(x@tax_table)[,7]

dff = data.frame(x@otu_table) %>% 
  mutate_if(is.numeric, ~1 * (. > 0))

df.t = data.frame(t(dff))

first = data.frame(df.t) %>% 
  select(matches("_first"))
colnames(first) = sub("X", "", colnames(first))
colnames(first) = sub("_first", "", colnames(first))

second = data.frame(df.t) %>% 
  select(matches("_second"))
colnames(second) = sub("X", "", colnames(second))
colnames(second) = sub("_second", "", colnames(second))

test = data.frame(first+second)
colnames(test) = sub("X", "", colnames(test))
#


table(test==0) # 14 498 detections total
table(test==1) # 8402 = 1 ##### 8402 / (8402+6096)
table(test==2) # 6096 = 2

c = test[rownames(test) %in% "Cryptostroma_corticale", ]
table(c==0) 
table(c==1) 
table(c==2)

c = test[rownames(test) %in% "Mutinus_ravenelii", ]
table(c==0) 
table(c==1) 
table(c==2)
#




fff = full_join(ff, fsec)

recapture = fff %>% 
  distinct(species); recapture



```


2. fold increase:

```{r}
twice = subset_samples(fungi, interval == "double"); twice = prune_taxa(taxa_sums(twice)>0,
                                                                         twice); twice
first = subset_samples(twice, grepl(pattern = "_1", 
                                    x = twice@sam_data$sample_id)); first = prune_taxa(taxa_sums(first)>0,
                                                                                       first); first
first@sam_data$turn = "first"; head(first@sam_data) # add a variable to distinguish between first and merged samples
second = merge_samples(twice, group = "pot_id"); second 

# reassign sample data in second:
second@sam_data$pot_id = rownames(second@sam_data) # fix pot_id variable
second@sam_data$pot_id = sub("p","", second@sam_data$pot_id) 
second@sam_data$pot_id = as.numeric(second@sam_data$pot_id) # this will let us sort by pot_id
second.sam = data.frame(second@sam_data) %>%  # using arrange() in dplyr
  arrange(pot_id) %>%
  mutate(pot_id=paste0("p",pot_id)); head(second.sam); tail(second.sam)



# now make an identical dataframe to get variables from: 
sam = data.frame(twice@sam_data) 
sam = sam %>%
  filter(grepl(pattern = "_2", 
               sam$sample_id))

# and reassign variables: 
second.sam$host = sam$host
second.sam$container = sam$container
second.sam$interval = sam$interval
second.sam$turn = "second"; # add a variable to distinguish between merged and first samples

# convert til sample_data() object:
second.sam = sample_data(second.sam); sample_names(second.sam)

# and reassign to twice.m phyloseq object: 
sample_data(second) = second.sam; twice.m; head(second@sam_data)


# now we can merge the two phyloseq objects in order to normalize them:
fiphy = merge_phyloseq(first, second); fiphy

# normalize the new phyloseq object:
sort(sample_sums(fiphy)) # we can normalize at the lowest read depth to keep all samples and only lose two taxa:
fiphyrar = rarefy_even_depth(fiphy, 
                             rngseed = 123, 
                             replace = FALSE, 
                             trimOTUs = TRUE, 
                             sample.size = 213969); fiphy; fiphyrar

# now we have to separate the new object into two dataframes before we combine
# them to calculate and plot fold increase: 
first = subset_samples(fiphyrar, turn=="first")
first.df = data.frame(cbind(estimate_richness(first, 
                                              measures = "Observed"),
                            first@sam_data)) %>% 
  rename("obs_first"="Observed",
         "pot_id_first"="pot_id",
         "turn_first"="turn") %>% 
  mutate(pot_id=pot_id_first); head(first.df)

second = subset_samples(fiphyrar, turn=="second")
second.df = data.frame(cbind(estimate_richness(second, 
                                              measures = "Observed"),
                            second@sam_data)) %>% 
  rename("obs_second"="Observed",
         "pot_id_second"="pot_id",
         "turn_second"="turn") %>% 
  mutate(pot_id=pot_id_second); head(second.df)

fidf = full_join(first.df, second.df, by = "pot_id") %>% 
  select("obs_first", "obs_second", 
         "pot_id", "pot_id_first", "pot_id_second", 
         "turn_first", "turn_second")

fidf = fidf %>% 
  mutate(foldincrease = obs_second/obs_first)

fig3.b.f=ggplot(fidf, aes(x=reorder(pot_id, -foldincrease), 
                 y=foldincrease,
                 fill="tan1", color="black")) + 
  geom_col(width=1) +
  scale_fill_manual(values = "tan1") +
  scale_color_manual(values = "black") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.ticks.x = element_blank()
      ) + 
  ylab("Fold increase") + 
  xlab("Sampled pots") +
  scale_x_discrete(labels = paste0(rep(" ", times=20))) +
  labs(title = " ",
       subtitle = "Fungi") +
  annotate(geom = "text", 
           x=5, 
           y=1.75, 
           label = paste0("Median = ", 
                         round(median(fidf$foldincrease), 
                               digits = 2)), # calculate and annotate manually...
           fontface = "italic", 
           size=3, 
           color="black", 
           hjust=0) +
  scale_y_continuous(breaks = seq(0, 3, 0.5), limits = c(0, 3)) + 
  coord_cartesian(ylim = c(0.9, 3)); fig3.b.f

# additionally, we will investigate whether there is a correlation between 
# the richness in the second sample and fold increase for each pot:
cormod = cor.test(fidf$foldincrease, fidf$obs_second, 
         method = "pearson")
fig3.c.f=ggplot(fidf, aes(x=obs_second, 
                 y=foldincrease,
                 fill="tan1")) + 
  theme_classic() +
  theme(legend.position = "none", 
        ) +
  scale_color_manual(values = "black") + 
  scale_fill_manual(values = "tan1") +
  geom_point(size=6, pch=21) + 
  geom_smooth(method = "lm", 
              color="black", 
              fill="grey90") +
  #scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 4, 0.5), 
                    limits = c(0, 3.6)) +
  xlab("Richness in both samples") +
  ylab("\nFold increase") + 
  labs(title = " ",
       subtitle = "Fungi") +
  coord_cartesian(ylim = c(1,3)) + 
  annotate(geom = "text", 
           x=750, 
           y=2.5, 
           label = paste0("r = ", 
                          round(cormod$estimate, 
                                digits = 3),
                          "\nP = ",
                          round(cormod$p.value,
                                digits = 3)), 
           fontface = "italic", size=3, color="black", 
           hjust = 0); fig3.c.f


```


combine into Fig. 3:

```{r}
fig3 = ggarrange(fig3.a.p,NULL,fig3.b.p,NULL,fig3.c.p,
                 fig3.a.f,NULL,fig3.b.f,NULL,fig3.c.f,
                 nrow=2,ncol=5, 
                 widths = c(3.5, 0.4, 3.5, 0.4, 3.5,
                            3.5, 0.4, 3.5, 0.4, 3.5)); fig3 # pdf landscape 6 x 10
```



